---
title: "Non Standard Evaluation in dplyr with Galaaz"
author:
    - "Rodrigo Botafogo"
    - "Daniel Mossé - University of Pittsburgh"
tags: [Tech, Data Science, Ruby, R, GraalVM]
date: "10/05/2019"
output:
  html_document:
    self_contained: true
    keep_md: true
  pdf_document:
    includes:
      in_header: ["../../sty/galaaz.sty"]
    number_sections: yes
    toc: true
    toc_depth: 2
  md_document:
    variant: markdown_github
fontsize: 11pt
---

```{r setup, echo=FALSE, message = FALSE}
library('dplyr')
library('tibble')
```

# Introduction

According to Steven Sagaert answer on Quora about "Is programming language R overrated?":

> R is a sophisticated language with an unusual (i.e. non-mainstream) set of features. It‘s 
> an impure functional programming language with sophisticated metaprogramming and 3 
> different OO systems.

> Just like common lisp you can completely customise how things work via metaprogramming. 
> The biggest example is the tidyverse: by creating it’s own evaluation system (tidyeval)
> was able to create a custom syntax for dplyr.

> Mastering R (the language) and its ecosystem is not a matter of weeks or months but 
> takes years. The rabbit hole goes pretty deep…

Although having a highly configurable language might give extreme power to the programmer,
it can also be, as stated above, a question of years to master it. Programming with _dplyr_
for instance, requires learning a set of complex concepts and rules that are not easily 
accessible for casual users or _unsofisticated_ programmers as many users of R are. Being
_unsofisticated_ is NOT used here in a negative sense, as R was build for statitians and
not programmers, that need to solve real problems, often in a short time spam and are not
concerned about creating complex computer systems.

Unfortunatelly, if this _unsofisticated_ programmer decides to move unto more sofisticated
coding, the learning curve might become a serious impediment. 

In this post we will see how to program with _dplyr_ in Galaaz and how Ruby can simplify
the learning curve of mastering _dplyr_ coding.

# But first, what is Galaaz??

Galaaz is a system for tightly coupling Ruby and R.  Ruby is a powerful language, with
a large community, a very large set of libraries and great for web development.  It is also
easy to learn. However,
it lacks libraries for data science, statistics, scientific plotting and machine learning.
On the other hand, R is considered one of the most powerful languages for solving all of the
above problems.  Maybe the strongest competitor to R is Python with libraries such as NumPy,
Pandas, SciPy, SciKit-Learn and many more.  We will not get here in the discussion on R
versus Python, both are excellent languages with powerful features, benefits and drawbacks.
Our interest is to bring to yet another excellent language, Ruby, the data science libraries
that it lacks.

With Galaaz we do not intend to re-implement any of the scientific libraries in R. However, we 
allow for very tight coupling between the two languages to the point that the Ruby 
developer does not need to know that there is an R engine running.  Also, from the point of
view of the R user/developer, Galaaz looks a lot like R, with just minor syntactic difference,
so there is almost no learning curve for the R developer. And as we will see in this
post that programming with _dplyr_ is easier in Galaaz than in R.

R users are probably quite knowledgeable about _dplyr_. For the Ruby developer, _dplyr_ and
the _tidyverse_ libraries are a set of libraries for data manipulation in R, developed by
Hardley Wickham, chief scientis at RStudio and a prolific R coder and writer.

For the coupling of Ruby and R, we use new technologies provided by Oracle: GraalVM, 
TruffleRuby and FastR. GraalVM home page had the following definition:

     GraalVM is a universal virtual machine for running applications
     written in JavaScript, Python 3, Ruby, R, JVM-based languages like Java,
     Scala, Kotlin, and LLVM-based languages such as C and C++.

     GraalVM removes the isolation between programming languages and enables
     interoperability in a shared runtime. It can run either standalone or in
     the context of OpenJDK, Node.js, Oracle Database, or MySQL.

     GraalVM allows you to write polyglot applications with a seamless way to
     pass values from one language to another. With GraalVM there is no copying
     or marshaling necessary as it is with other polyglot systems. This lets
     you achieve high performance when language boundaries are crossed. Most
     of the time there is no additional cost for crossing a language boundary
     at all.

     Often developers have to make uncomfortable compromises that require them
     to rewrite their software in other languages. For example:

      * “That library is not available in my language. I need to rewrite it.” 
      * “That language would be the perfect fit for my problem, but we cannot
        run it in our environment.” 
      * “That problem is already solved in my language, but the language is
        too slow.”
    
    With GraalVM we aim to allow developers to freely choose the right language
    for the task at hand without making compromises.


# Tidyverse and dplyr

In [What is the tidyverse?](https://rviews.rstudio.com/2017/06/08/what-is-the-tidyverse/) the 
tidyverse is explained as follows:

> The tidyverse is a coherent system of packages for data manipulation, exploration and 
> visualization that share a common design philosophy. These were mostly developed by 
> Hadley Wickham himself, but they are now being expanded by several contributors. Tidyverse 
> packages are intended to make statisticians and data scientists more productive by 
> guiding them through workflows that facilitate communication, and result in reproducible 
> work products. Fundamentally, the tidyverse is about the connections between the tools 
> that make the workflow possible.

_dplyr_ is one of the many packages that are part of the tidyverse.  It is:

> a grammar of data manipulation, providing a consistent set of verbs that help you solve 
> the most common data manipulation challenges:

> 1. mutate() adds new variables that are functions of existing variables
> 2. select() picks variables based on their names.
> 3. filter() picks cases based on their values.
> 4. summarise() reduces multiple values down to a single summary.
> 5. arrange() changes the ordering of the rows.

Very often R is used interactively and users use _dplyr_ to manipulate a single dataset
without programming. When users want to replicate their work for
multiple datasets, programming becomes necessary.

# Programming with dplyr

In the vignette ["Programming with dplyr"](https://dplyr.tidyverse.org/articles/programming.html), 
Hardley Wickham states:

> Most dplyr functions use non-standard evaluation (NSE). This is a catch-all term that 
> means they don’t follow the usual R rules of evaluation. Instead, they capture the 
> expression that you typed and evaluate it in a custom way. This has two main 
> benefits for dplyr code:

> Operations on data frames can be expressed succinctly because you don’t need to repeat 
> the name of the data frame. For example, you can write filter(df, x == 1, y == 2, z == 3) 
> instead of df[df\$x == 1 & df\$y ==2 & df\$z == 3, ].

> dplyr can choose to compute results in a different way to base R. This is important for 
> database backends because dplyr itself doesn’t do any work, but instead generates the SQL 
> that tells the database what to do.

But then he goes on:

> Unfortunately these benefits do not come for free. There are two main drawbacks:

> Most dplyr arguments are not referentially transparent. That means you can’t replace a value 
> with a seemingly equivalent object that you’ve defined elsewhere. In other words, this code:

```{r tibble, eval=FALSE}
df <- data.frame(x = 1:3, y = 3:1)
print(filter(df, x == 1))
#> # A tibble: 1 x 2
#>       x     y
#>   <int> <int>
#> 1     1     3
```
> Is not equivalent to this code:

```{r my_var_err, eval = FALSE}
my_var <- x
#> Error in eval(expr, envir, enclos): object 'x' not found
filter(df, my_var == 1)
#> Error: object 'my_var' not found
```
> This makes it hard to create functions with arguments that change how dplyr verbs are computed.

As a result of this, programming with _dplyr_ requires learning a set of new ideas and concepts.
In this vignette Hardley goes on showing how to program ever more difficult problems with _dplyr_, 
showing the problems it faces and the new concepts needed to solve them.

In this blog, we will look at all the problems presented by Harley on the vignette and show how
those same problems can be solved using Galaaz and the Ruby language. 

This blog is organized as follows: first we show how to write expressions using Galaaz.  
Expressions are a fundamental concept in _dplyr_ and are not part of basic Ruby.  We extend
the Ruby language create a manipulate expressions that will be used by _dplyr_ functions. 

Then we show very succintly how Ruby and R can be integrated and how R functions are 
transparently called from Ruby. Galaaz [user manual](https://github.com/rbotafogo/galaaz/wiki)
(still in development) goes in much deeper detail about this integration.

Next in section "Data manipulation wiht _dplyr_" we go through all the problems on the 
_dplyr_ vignette and look at how they are solved in Galaaz.  We then discuss why programming
with Galaaz and _dplyr_ is easier than programming with _dplyr_ in plain R.

The following section looks at another more advanced problem and shows that Galaaz can still
handle it without any difficulty.  We then provide further reading and concluding remarks.

# Writing Expressions in Galaaz

Galaaz extends Ruby to work with expressions, similar to R's expressions build with 'quote' 
(base R) or 'quo' (tidyverse).  Expressions in this context are like mathematical expressions or
formulae.  For instance, in mathematics, the expression $y = sin(x)$ describes a function but cannot
be computed unless the value of $x$ is bound to some value. 

Expressions are fundamental in _dplyr_ programming as they are the input to _dplyr_ functions,
for instance, as we will see shortly, if a data frame has a column named 'x' and we want 
to add another column, y, to this dataframe that has the values of 'x' times 2, then we would
call a _dplyr_ function with the expression 'y = x * 2'.

## A note on notation

This blog was written in Rmarkdown and automatically converted to HTML or PDF (depending on
where you are reading this blog) with gKnit (a tool provided by Galaaz). In Rmarkdown, it is
possible to write text and code blocks that are executed to generate the final report. Code
blocks appear inside a 'box' and the result of their execution appear either in another type 
of 'box' with a different background (HTML) or as normal text (PDF).  Every output line from
the code execution is preceeded by '##'.

## Expressions from operators

The code below creates an expression summing two symbols. Note that :a and :b are Ruby symbols and
are not bound to any values at the time of expression definition:

```{ruby expressions}
exp1 = :a + :b
puts exp1
```
In Galaaz, we can build any complex mathematical expression such as:

```{ruby expr2}
exp2 = (:a + :b) * 2.0 + :c ** 2 / :z
puts exp2
```
Expressions are printed with the same format as the equivalent R expressions. The 'L' after 
2 indicates that 2 is an integer. 

The R developer should note that in R, if she writes the
number '2', the R interpreter will convert it to float. In order to get an interger she
should write '2L'. Galaaz follows Ruby notation and '2' is an integer, while '2.0' is a
float.

It is also possible to use inequality operators in building expressions:

```{ruby expr3}
exp3 = (:a + :b) >= :z
puts exp3
```
Expressions' definition can also make use of normal Ruby variables without any problem:

```{ruby expr_with_var}
x = 20
y = 30.0
exp_var = (:a + :b) * x <= :z - y
puts exp_var
```

Galaaz provides both symbolic representations for operators, such as (>, <, !=) as functional 
notation for those operators such as (.gt, .ge, etc.).  So the same expression written 
above can also be written as

```{ruby expr4}
exp4 = (:a + :b).ge :z
puts exp4
```

Two types of expressions, however, can only be created with the functional representation 
of the operators.  Those are expressions involving '==', and '='. This is the case since
those symbols have special meaning in Ruby and should not be redefined.

In order to write an expression involving '==' we
need to use the method '.eq' and for '=' we need the function '.assign':

```{ruby expr5}
exp5 = (:a + :b).eq :z
puts exp5
```

```{ruby expr6}
exp6 = :y.assign :a + :b
puts exp6
```
Users should be careful when writing expressions not to inadvertently use '==' or '=' as 
this will generate an error, that might be a bit cryptic (in future releases of Galaza, we 
plan to improve the error message).

```{ruby exp_wrong, warning=FALSE}
exp_wrong = (:a + :b) == :z
puts exp_wrong
```
The problem lies with the fact that
when using '==' we are comparing expression (:a + :b) to expression :z with '=='.  When this 
comparison is executed, the system tries to evaluate :a, :b and :z, and those symbols, at 
this time, are not bound to anything giving the "object 'a' not found" message.

## Expressions with R methods

It is often necessary to create an expression that uses a method or function.  For instance, in
mathematics, it's quite natural to write an expressin such as $y = sin(x)$. In this case, the
'sin' function is part of the expression and should not be immediately executed. When we want 
the function to be part of the expression, we call the function preceeding it
by the letter E, such as 'E.sin(x)'

```{ruby method_expression}
exp7 = :y.assign E.sin(:x)
puts exp7
```
Function expressions can also be written using '.' notation:

```{ruby expression_with_dot}
exp8 = :y.assign :x.sin
puts exp8
```
When a function has multiple arguments, the first one can be used before the '.'.  For instance,
the R concatenate function 'c', that concatenates two or more arguments can be part of 
an expression as:

```{ruby expression_multiple_args}
exp9 = :x.c(:y)
puts exp9
```
Note that this gives an OO feeling to the code, as if we were saying 'x' concatenates 'y'. As a
side note, '.' notation can be used as the R pipe operator '%>%', but is more general than the
pipe.

## Evaluating an Expression

Although we are mainly focusing on expressions to pass them to _dplyr_ functions, expressions 
can be evaluated by calling function 'eval' with a binding. 

A binding can be provided with a list or a data frame as shown below:

```{ruby eval_expression_list}
exp = (:a + :b) * 2.0 + :c ** 2 / :z
puts exp.eval(R.list(a: 10, b: 20, c: 30, z: 40))
```

with a data frame:

```{ruby eval_expression_df}
df = R.data__frame(
  a: R.c(1, 2, 3),
  b: R.c(10, 20, 30),
  c: R.c(100, 200, 300),
  z: R.c(1000, 2000, 3000))

puts exp.eval(df)
```

# Using Galaaz to call R functions

Galaaz tries to emulate as closely as possible the way R functions are called and migrating from 
R to Galaaz should be quite easy requiring only minor syntactic changes to an R script.  In 
this post, we do not have enough space to write a complete manual on Galaaz 
(a short manual can be found at: https://www.rubydoc.info/gems/galaaz/0.4.9), so we will 
present only a few examples scripts using Galaaz.

Basically, to call an R function from Ruby with Galaaz, one only needs to preced the function
with 'R.'.  For instance, to create a vector in R, the 'c' function is used. In Galaaz, a
vector can be created by using 'R.c':

```{ruby vector}
vec = R.c(1.0, 2, 3)
puts vec
```
A list is created in R with the 'list' function, so in Galaaz we do:

```{ruby list}
list = R.list(a: 1.0, b: 2, c: 3)
puts list
```
Note that we can use named arguments in our list.  The same code in R would be:

```{r list2}
lst = list(a = 1, b = 2L, c = 3L)
print(lst)
```
Now, let's say that 'x' is an angle of 45$^\circ$ and we acttually want to create
the expression $y = sin(45^\circ)$, which is $y = 0.850...$.  In this case, 
we will use 'R.sin':

```{ruby eval_sin}
exp10 = :y.assign R.sin(45)
puts exp10
```

# Data manipulation wiht _dplyr_

In this section we will give a brief tour _dplyr_'s usage in Galaaz and how to manipulate 
data in Ruby with it.  This section will follow [_dplyr_'s vignette](https://dplyr.tidyverse.org/articles/dplyr.html) that explores the nycflights13 data set. This dataset contains all 336776 
flights that departed from New York City in 2013. The data comes from the US Bureau of 
Transportation Statistics.

Let's start by taking a look at this dataset:

```{ruby nycflights13}
R.library('nycflights13')
# check it's dimension
puts ~:flights.dim
# and the structure
~:flights.str
```

Now, let's use a first verb of _dplyr_: 'filter'. This verb, obviously, will filter the data
by the given expression. In the next block, we filter by columns 'month' and 'day'. The
first argument to the filter function is symbol ':flights'. A Ruby symbol, when given to
an R function will convert to the R variable of the same name, in this case 'flights', that
holds the nycflights13 data frame. 

The second and third arguments are expressions that will be used by the filter function to
filter by columns, looking for entries in which the month and day are equal to 1.

```{ruby filter}
puts R.filter(:flights, (:month.eq 1), (:day.eq 1))
```


## Programming with _dplyr_: problems and how to solve them in Galaaz

In this section we look at the list of problems that Hardley describes in the "Programming with dplyr"
vignette and show how those problems are solved and coded with Galaaz. Readers interested in
how those problems are treated in _dplyr_ should read the vignette and use it as a comparison with
this blog.

## Filtering using expressions

Now that we know how to write expressions and call R functions, let's do some data manipulation in
Galaaz.  Let's first start by creating a data frame. In R, the 'data.frame' function creates a
data frame.  In Ruby, writing 'data.frame' will not parse as a single object. To call R 
functions that have a '.' in them, we need to substitute the '.' with '__'. So, method
'data.frame' in R, is called in Galaaz as 'R.data\_\_frame':

```{ruby df}
df = R.data__frame(x: (1..3), y: (3..1))
puts df
```

_dplyr_ provides the 'filter' function, that filters data in a data brame. The 'filter' 
function can be called on this data frame either by using 'R.filter(df, ...)' or 
by using dot notation.  

-------FIX---------

We prefer to use dot notation as shown bellow.  The argument to 'filter' should be an 
expression. Note that if we gave to filter a Ruby expression such as
'x == 1', we would get an error, since there is no variable 'x' defined and if 'x' was a variable
then 'x == 1' would either be 'true' or 'false'. Our goal is to filter our data frame returning
all rows in which the 'x' value is equal to 1. To express this we want: ':x.eq 1', where :x will
be interpreted by filter as the 'x' column.

```{ruby filter_exp}
puts df.filter(:x.eq 1)
```
In R, and when coding with 'tidyverse', arguments to a function are usually not 
*referencially transparent*. That is, you can’t replace a value with a seemingly equivalent 
object that you’ve defined elsewhere. In other words, this code

```{r not_transp, eval=FALSE}
my_var <- x
filter(df, my_var == 1)
```
Generates the following error: "object 'x' not found.

However, in Galaaz, arguments are referencially transparent as can be seen by the 
code bellow.  Note initally that 'my_var = :x' will not give the error "object 'x' not found" 
since ':x' is treated as an expression and assigned to my\_var. Then when doing (my\_var.eq 1), 
my\_var is a variable that resolves to ':x' and it becomes equivalent to (:x.eq 1) which is
what we want.

```{ruby my_var}
my_var = :x
puts df.filter(my_var.eq 1)
```
As stated by Hardley

> dplyr code is ambiguous. Depending on what variables are defined where, 
> filter(df, x == y) could be equivalent to any of:

```
df[df$x == df$y, ]
df[df$x == y, ]
df[x == df$y, ]
df[x == y, ]
```
In galaaz this ambiguity does not exist, filter(df, x.eq y) is not a valid expression as 
expressions are build with symbols.  In doing filter(df, :x.eq y) we are looking for elements
of the 'x' column that are equal to a previously defined y variable.  Finally in 
filter(df, :x.eq :y) we are looking for elements in which the 'x' column value is equal to
the 'y' column value. This can be seen in the following two chunks of code:

```{ruby disamb1}
y = 1
x = 2

# looking for values where the 'x' column is equal to the 'y' column
puts df.filter(:x.eq :y)
```

```{ruby disamb2}
# looking for values where the 'x' column is equal to the 'y' variable
# in this case, the number 1
puts df.filter(:x.eq y)
```
## Writing a function that applies to different data sets

Let's suppose that we want to write a function that receives as the first argument a data frame
and as second argument an expression that adds a column to the data frame that is equal to the
sum of elements in column 'a' plus 'x'. 

Here is the intended behaviour using the 'mutate' function of 'dplyr':

```
mutate(df1, y = a + x)
mutate(df2, y = a + x)
mutate(df3, y = a + x)
mutate(df4, y = a + x)
```
The naive approach to writing an R function to solve this problem is:

```
mutate_y <- function(df) {
  mutate(df, y = a + x)
}
```
Unfortunately, in R, this function can fail silently if one of the variables isn’t present 
in the data frame, but is present in the global environment.  We will not go through here how
to solve this problem in R.

In Galaaz the method mutate_y bellow will work fine and will never fail silently.

```{ruby mutate_y, warning=FALSE}
def mutate_y(df)
  df.mutate(:y.assign :a + :x)
end
```
Here we create a data frame that has only one column named 'x':

```{ruby data_frame_no_a_column, warning=FALSE}
df1 = R.data__frame(x: (1..3))
puts df1
```

Note that method mutate_y will fail independetly from the fact that variable 'a' is defined and
in the scope of the method.  Variable 'a' has no relationship with the symbol ':a' used in the
definition of 'mutate\_y' above:

```{ruby call_mutate_y, warning = FALSE}
a = 10
mutate_y(df1)
```
## Different expressions

Let's move to the next problem as presented by Hardley where trying to write a function in R 
that will receive two argumens, the first a variable and the second an expression is not trivial.
Bellow we create a data frame and we want to write a function that groups data by a variable and
summarises it by an expression:

```{r diff_expr}
set.seed(123)

df <- data.frame(
  g1 = c(1, 1, 2, 2, 2),
  g2 = c(1, 2, 1, 2, 1),
  a = sample(5),
  b = sample(5)
)

as.data.frame(df) 

d2 <- df %>%
  group_by(g1) %>%
  summarise(a = mean(a))
	       
as.data.frame(d2)	       

d2 <- df %>%
  group_by(g2) %>%
  summarise(a = mean(a))
	       
as.data.frame(d2)	       
```

As shown by Hardley, one might expect this function to do the trick:

```{r diff_exp_fnc}
my_summarise <- function(df, group_var) {
  df %>%
    group_by(group_var) %>%
    summarise(a = mean(a))
}

# my_summarise(df, g1)
#> Error: Column `group_var` is unknown
```

In order to solve this problem, coding with dplyr requires the introduction of many new concepts
and functions such as 'quo', 'quos', 'enquo', 'enquos', '!!' (bang bang), '!!!' (triple bang). 
Again, we'll leave to Hardley the explanation on how to use all those functions.

Now, let's try to implement the same function in galaaz.  The next code block first prints the
'df' data frame define previously in R (to access an R variable from Galaaz, we use the tilda 
operator '~' applied to the R variable name as symbol, i.e., ':df'.  We then create the 
'my_summarize' method and call it passing the R data frame and the group by variable ':g1':

```{ruby diff_exp_ruby_func}
puts ~:df
print "\n"

def my_summarize(df, group_var)
  df.group_by(group_var).
    summarize(a: :a.mean)
end

puts my_summarize(:df, :g1).as__data__frame
```
It works!!! Well, let's make sure this was not just some coincidence

```{ruby group_g2}
puts my_summarize(:df, :g2).as__data__frame
```

Great, everything is fine! No magic, no new functions, no complexities, just normal, standard Ruby
code.  If you've ever done NSE in R, this certainly feels much safer and easy to implement.

## Different input variables

In the previous section we've managed to get rid of all NSE formulation for a simple example, but
does this remain true for more complex examples, or will the Galaaz way prove inpractical for
more complex code?

In the next example Hardley proposes us to write a function that given an expression such as 'a'
or 'a * b', calculates three summaries.  What we want a function that does the same as these R
statements:

```
summarise(df, mean = mean(a), sum = sum(a), n = n())
#> # A tibble: 1 x 3
#>    mean   sum     n
#>   <dbl> <int> <int>
#> 1     3    15     5

summarise(df, mean = mean(a * b), sum = sum(a * b), n = n())
#> # A tibble: 1 x 3
#>    mean   sum     n
#>   <dbl> <int> <int>
#> 1   9    45     5
```

Let's try it in galaaz:

```{ruby summarize_method}
def my_summarise2(df, expr)
  df.summarize(
    mean: E.mean(expr),
    sum: E.sum(expr),
    n: E.n
  )
end

puts my_summarise2((~:df), :a)
puts my_summarise2((~:df), :a * :b)
```

Once again, there is no need to use any special theory or functions.  The only point to be 
careful about is the use of 'E' to build expressions from functions 'mean', 'sum' and 'n'.

## Different input and output variable

Now the next challenge presented by Hardley is to vary the name of the output variables based on 
the received expression.  So, if the input expression is 'a', we want our data frame columns to
be named 'mean\_a' and 'sum\_a'.  Now, if the input expression is 'b', columns
should be named 'mean\_b' and 'sum\_b'.

```
mutate(df, mean_a = mean(a), sum_a = sum(a))
#> # A tibble: 5 x 6
#>      g1    g2     a     b mean_a sum_a
#>   <dbl> <dbl> <int> <int>  <dbl> <int>
#> 1     1     1     1     3      3    15
#> 2     1     2     4     2      3    15
#> 3     2     1     2     1      3    15
#> 4     2     2     5     4      3    15
#> # … with 1 more row

mutate(df, mean_b = mean(b), sum_b = sum(b))
#> # A tibble: 5 x 6
#>      g1    g2     a     b mean_b sum_b
#>   <dbl> <dbl> <int> <int>  <dbl> <int>
#> 1     1     1     1     3      3    15
#> 2     1     2     4     2      3    15
#> 3     2     1     2     1      3    15
#> 4     2     2     5     4      3    15
#> # … with 1 more row
```
In order to solve this problem in R, Hardley needs to introduce some more new functions and notations:
'quo_name' and the ':=' operator from package 'rlang'

Here is our Ruby code:

```{ruby name_change}
def my_mutate(df, expr)
  mean_name = "mean_#{expr.to_s}"
  sum_name = "sum_#{expr.to_s}"

  df.mutate(mean_name => E.mean(expr),
            sum_name => E.sum(expr))
end

puts my_mutate((~:df), :a)
puts my_mutate((~:df), :b)
```
It really seems that "Non Standard Evaluation" is actually quite standard in Galaaz! But, you 
might have noticed a small change in the way the arguments to the mutate method were called.
In a previous example we used df.summarise(mean: E.mean(:a), ...) where the column name was
followed by a ':' colom.  In this example, we have df.mutate(mean_name => E.mean(expr), ...)
and variable mean\_name is not followed by ':' but by '=>'.  This is standard Ruby notation.

[explain....]

## Capturing multiple variables

Moving on with new complexities, Hardley proposes us to solve the problem in which the 
summarise function will receive any number of grouping variables.

This again is quite standard Ruby.  In order to receive an undefined number of paramenters
the paramenter is preceded by '*':

```{ruby multiple_vars}
def my_summarise3(df, *group_vars)
  df.group_by(*group_vars).
    summarise(a: E.mean(:a))
end

puts my_summarise3((~:df), :g1, :g2).as__data__frame
```

# Why does R require NSE and Galaaz does not?

NSE introduces a number of new concepts, such as 'quoting', 'quasiquotation', 'unquoting' and 
'unquote-splicing', while in Galaaz none of those concepts are needed. What gives?

R is an extremely flexible language and it has lazy evaluation of parameters. When in R a 
function is called as 'summarise(df, a = b)', the summarise function receives the litteral
'a = b' parameter and can work with this as if it were a string. In R, it is not clear what
a and b are, they can be expressions or they can be variables, it is up to the function to
decide what 'a = b' means.

In Ruby, there is no lazy evaluation of parameters and 'a' is always a variable and so is 'b'.
Variables assume their value as soon as they are used, so 'x = a' is immediately evaluate and 
variable 'x' will receive the value of variable 'a' as soon as the Ruby statement is executed. 
Ruby also provides the notion of a symbol; ':a' is a symbol and does not evaluate to anything.
Galaaz uses Ruby symbols to build expressions that are not bound to anything: ':a.eq :b' is
clearly an expression and has no relationship whatsoever with the statment 'a = b'. By using
symbols, variables and expressions all the possible ambiguities that are found in R are 
eliminated in Galaaz.

The main problem that remains, is that in R, functions are not clearly documented as what type
of input they are expecting, they might be expecting regular variables or they might be 
expecting expressions and the R function will know how to deal with an input of the form 
'a = b', now for the Ruby developer it might not be immediately clear if it should call the 
function passing the value 'true' if variable 'a' is equal to variable 'b' or if it should
call the function passing the expression ':a.eq :b'.


# Advanced dplyr features

In the blog: [Programming with dplyr by using dplyr](https://www.r-bloggers.com/programming-with-dplyr-by-using-dplyr/) Iñaki Úcar shows surprise that some R users are trying to code in dplyr avoiding
the use of NSE.  For instance he says:

> Take the example of seplyr. It stands for standard evaluation dplyr, and enables us to 
> program over dplyr without having “to bring in (or study) any deep-theory or 
> heavy-weight tools such as rlang/tidyeval”.

For me, there isn't really any surprise that users are trying to avoid dplyr deep-theory. R
users frequently are not programmers and learning to code is already hard business, on top
of that, having to learn how to 'quote' or 'enquo' or 'quos' or 'enquos' is not necessarily
a 'piece of cake'. So much so, that 'tidyeval' has some more advanced functions that instead
of using quoted expressions, uses strings as arguments.

In the following examples, we show the use of functions 'group\_by\_at', 'summarise\_at' and
'rename\_at' that receive strings as argument. The data frame used in 'starwars' that describes
features of characters in the Starwars movies:

```{ruby starwars}
puts (~:starwars).head.as__data__frame
```
The grouped_mean function bellow will receive a grouping variable and calculate summaries for
the value\_variables given:

```{r grouped_mean}
grouped_mean <- function(data, grouping_variables, value_variables) {
  data %>%
    group_by_at(grouping_variables) %>%
    mutate(count = n()) %>%
    summarise_at(c(value_variables, "count"), mean, na.rm = TRUE) %>%
    rename_at(value_variables, funs(paste0("mean_", .)))
    }

gm = starwars %>% 
   grouped_mean("eye_color", c("mass", "birth_year"))

as.data.frame(gm)   
```

The same code with Galaaz, becomes:

```{ruby advanced_starwars}
def grouped_mean(data, grouping_variables, value_variables)
  data.
    group_by_at(grouping_variables).
    mutate(count: E.n).
    summarise_at(E.c(value_variables, "count"), ~:mean, na__rm: true).
    rename_at(value_variables, E.funs(E.paste0("mean_", value_variables)))
end

puts grouped_mean((~:starwars), "eye_color", E.c("mass", "birth_year")).as__data__frame
```

# Further reading

For more information on GraalVM, TruffleRuby, fastR, R and Galaaz check out the following sites/posts:

* [GraalVM Home](https://www.graalvm.org/)
* [TruffleRuby](https://github.com/oracle/truffleruby)
* [FastR](https://github.com/oracle/fastr)
* [Faster R with FastR](https://medium.com/graalvm/faster-r-with-fastr-4b8db0e0dceb)
* [How to make Beautiful Ruby Plots with Galaaz](https://medium.freecodecamp.org/how-to-make-beautiful-ruby-plots-with-galaaz-320848058857)
* [Ruby Plotting with Galaaz: An example of tightly coupling Ruby and R in GraalVM](https://towardsdatascience.com/ruby-plotting-with-galaaz-an-example-of-tightly-coupling-ruby-and-r-in-graalvm-520b69e21021)
* [How to do reproducible research in Ruby with gKnit](https://towardsdatascience.com/how-to-do-reproducible-research-in-ruby-with-gknit-c26d2684d64e)
* [R for Data Science](https://r4ds.had.co.nz/)
* [Advanced R](https://adv-r.hadley.nz/)

# Conclusion

Ruby and Galaaz provide a nice framework for developing code that uses R functions. Although R is
a very powerful and flexible language, sometimes, too much flexibility makes life harder for
the casual user. We believe however, that even for the advanced user, Ruby integrated 
with R throught Galaaz, makes a powerful environment for data analysis.  In this blog post we
showed how Galaaz consistent syntax eliminates the need for complex constructs such as quoting,
enquoting, quasiquotation, etc. This simplification comes from the fact that expressions and
variables are clearly separated objects, which is not the case in the R language.
