---
title: "Galaaz Manual"
subtitle: "Coupling Ruby (JRuby) and GNU R for data science"
author: "Rodrigo Botafogo"
tags: [Galaaz, Ruby, JRuby, R, "GNU R", ggplot2, knitr, dplyr, Bioconductor, Arrow]
date: "2026"
bibliography: "../../examples/Bibliography/stats.bib"
output:
  html_document:
    self_contained: true
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    includes:
      in_header: "../../sty/galaaz.sty"
    keep_tex: yes
    number_sections: yes
    toc: true
    toc_depth: 3
  md_document:
    variant: markdown_github
fontsize: 11pt
---

```{ruby setup, echo=FALSE}
# Bridge default is 60s; some chunks (Arrow, large dplyr pipes) need more.
ENV['GALAAZ_BRIDGE_TIMEOUT_SEC'] ||= '300'

R.options(crayon__enabled: false)
R.install_and_loads('kableExtra')
```

# Introduction

Galaaz is a system for tightly coupling Ruby and R. Ruby is a powerful language, with a large 
community, a very large set of libraries and great for web development. However, it lacks 
libraries for data science, statistics, scientific plotting and machine learning. On the 
other hand, R is considered one of the most powerful languages for solving all of the above 
problems. **Python** is a strong competitor: NumPy, pandas, SciPy, and scikit-learn are
widely used building blocks, and **PyPI** hosts many thousands of other packages for
numerical work, machine learning, and beyond.

With Galaaz we do not intend to re-implement any of the scientific libraries in R, we allow
for very tight coupling between the two languages to the point that the Ruby developer does
not need to know that there is an R engine running.

According to Wikipedia "Ruby is a dynamic, interpreted, reflective, object-oriented, 
general-purpose programming language. It was designed and developed in the mid-1990s by Yukihiro 
"Matz" Matsumoto in Japan."  It reached high popularity with the development of Ruby on Rails
(RoR) by David Heinemeier Hansson. RoR is a web application framework first released
around 2005. It makes extensive use of Ruby's metaprogramming features.  With RoR,
Ruby became very popular.  According to [Ruby’s place in the TIOBE index](https://www.tiobe.com/tiobe-index/ruby/)
it peaked in popularity around 2008, then declined until 2015 when it started picking up again.
Ruby remains a significant language in web development and general-purpose scripting.

Python, a language similar to Ruby, ranks 4th in the index.  Java, C and C++ take the
first three positions.  Ruby is often criticized for its focus on web applications.
But Ruby can do [much more](https://github.com/markets/awesome-ruby) than just web applications.
Yet, for scientific computing, Ruby lags behind Python and R.  Python offers Django and
similar frameworks for the web, plus NumPy, pandas, and a deep catalog of science and ML libraries.
R is a free software environment for statistical computing and graphics with thousands
of libraries for data analysis. 

Until recently, there was no real perspective for Ruby to bridge this gap.
Implementing a complete scientific computing infrastructure would take too long.

**Galaaz 2.0** couples **JRuby** (Ruby on the JVM) with **GNU R**—the same R you use for
CRAN and Bioconductor. Ruby and R run in **separate processes**; the **Galaaz bridge**
sends requests to R and returns results to Ruby. From your point of view you still write
Ruby: `R.c(...)`, `R.library('ggplot2')`, `~R[:mtcars]`, and dplyr-style chains on R objects.
You do not need to learn R syntax to get a lot done, though reading R documentation for
individual packages remains useful.

Earlier experiments with Galaaz used Oracle’s **GraalVM** with TruffleRuby and FastR so that
Ruby and R could share one runtime. That path is no longer the focus: **standard GNU R**
gives full compatibility with the R package ecosystem (including compiled extensions and
Bioconductor) while JRuby gives a mature Ruby with **real multithreading** for application
and I/O code.

The bridge handles **communication and typing** between the two worlds; large tables can
also flow through **Apache Arrow** on the R side when you use the optional helpers described
later in this manual.

Library wrapping is a common way to bring features from one language into another.
To improve performance, Python often wraps more efficient C libraries. For the
Python developer, the existence of such C libraries is hidden.  The problem with
library wrapping is that for any new library, there is the need to handcraft a new
wrapper.

Galaaz, instead of wrapping a single C or R library, wraps the whole R language
in Ruby.  Doing so, all thousands of R libraries are available immediately
to Ruby developers without any new wrapping effort.

## What does Galaaz mean

Galaaz is the Portuguese name for "Galahad".  From Wikipedia:

    Sir Galahad (sometimes referred to as Galeas or Galath),
    in Arthurian legend, is a knight of King Arthur's Round Table and one
    of the three achievers of the Holy Grail. He is the illegitimate son
    of Sir Lancelot and Elaine of Corbenic, and is renowned for his
    gallantry and purity as the most perfect of all knights. Emerging quite
    late in the medieval Arthurian tradition, Sir Galahad first appears in the
    Lancelot–Grail cycle, and his story is taken up in later works such as
    the Post-Vulgate Cycle and Sir Thomas Malory's Le Morte d'Arthur.
    His name should not be mistaken with Galehaut, a different knight from
    Arthurian legend. 

# Command-line tools (`bin/`)

The Galaaz repository ships many helpers under **`bin/`**. When working from a **clone**, call
them as **`bin/<name>`** from the project root (or `./bin/<name>`). If you install the **gem**,
only a subset is guaranteed on your `PATH` (see the gemspec: **`galaaz`**, **`gstudio`**, **`gknit`**, **`grun`**, **`gknit-draft`**); for development and CI, prefer the **`bin/`** copies so JVM flags and paths stay correct.

Below, **current (Galaaz 2.0 + JRuby + GNU R)** means the tool is wired to **`jruby`** and
**`bin/galaaz_jruby_env.inc.sh`** (or equivalent logic in Ruby via `lib/galaaz_jruby.rb`). **Legacy**
means the script still targets **GraalVM** polyglot Ruby / FastR-era invocation and is **not**
expected to work on a typical JRuby-only setup.

**Table layout:** names in the first column are **`bin/`** filenames (run as `bin/<name>` from the repo root). Long options and examples sit **outside** the tables so PDF columns stay readable.

```{r bin-tables-helper, echo=FALSE}
bin_tbl <- function(df) {
  k <- knitr::kable(df, row.names = FALSE, booktabs = TRUE, linesep = "",
                    col.names = c("Script", "Role", "2.0?"))
  if (knitr::is_latex_output()) {
    k <- kableExtra::kable_styling(k, font_size = 9, latex_options = "scale_down")
    k <- kableExtra::column_spec(k, 1, width = "2.5cm")
    k <- kableExtra::column_spec(k, 2, width = "9.5cm")
    k <- kableExtra::column_spec(k, 3, width = "2.8cm")
  } else {
    k <- kableExtra::kable_styling(k, bootstrap_options = c("striped", "condensed"), full_width = TRUE)
  }
  k
}
```

```{r bin-tables-bootstrap, echo=FALSE}
df_boot <- data.frame(
  Script = c("galaaz-bootstrap", "galaaz-jruby", "galaaz_jruby_env.inc.sh", "install-tinytex"),
  Role = c(
    "WSL2 helper: Docker checks; optional TinyTeX or poppler for gKnit PDF.",
    "JRuby with repo lib/ on LOAD_PATH and required JVM flags (e.g. Arrow).",
    "Sourced by bash wrappers; sets GALAAZ_REQUIRED_JRUBY_J_ARGS.",
    "Install TinyTeX for PDF output."
  ),
  X2 = c("Yes*", "Yes", "Yes†", "Yes"),
  stringsAsFactors = FALSE
)
bin_tbl(df_boot)
```

\* Where WSL/Docker apply. **`galaaz-bootstrap` flags:** `--check`, `--apply`, `--runtime` (`docker` \| `local` \| `auto`), `--[no-]prompt-doc-tools`.

† Not run directly.

**`galaaz-jruby` examples** (from repo root):

```text
bin/galaaz-jruby my_script.rb
bin/galaaz-jruby -S rspec
```

## Interactive use, examples, and Rake

```{r bin-tables-interactive, echo=FALSE}
df_ix <- data.frame(
  Script = c("gstudio", "run_example", "galaaz"),
  Role = c(
    "IRB or Pry with Galaaz preloaded (JRuby + JVM flags).",
    "Run one Ruby file using the same JRuby/JVM setup as tests.",
    "Forward arguments to rake (needs rake; usually JRuby)."
  ),
  X2 = c("Yes", "Yes", "Yes"),
  stringsAsFactors = FALSE
)
bin_tbl(df_ix)
```

## gKnit and document drafts

```{r bin-tables-gknit, echo=FALSE}
df_gk <- data.frame(
  Script = c("gknit", "gknit-draft", "gknit-draft.rb", "gknit_Rscript"),
  Role = c(
    "Knit .Rmd via JRuby and R Markdown render.",
    "Drafts from rticles-style templates; wrapper still uses legacy polyglot ruby.",
    "Ruby entry: GKnit.draft (use with JRuby + LOAD_PATH).",
    "Polyglot Rscript launcher; hard-coded LOAD_PATH sample."
  ),
  X2 = c("Yes", "Legacy", "JRuby", "No"),
  stringsAsFactors = FALSE
)
bin_tbl(df_gk)
```

**`gknit` CLI** (see `gknit -h`): `--output_format`, `--output_file`, `--output_dir`, `--bridge_timeout_sec`, `--callback_timeout_ms`. If `--output_format` is omitted, the **first** YAML `output:` target wins.

Prefer **`galaaz-jruby`** for **`gknit-draft`** workflows until that wrapper matches the **`gknit`** stack.

## Tests

```{r bin-tables-tests, echo=FALSE}
df_ts <- data.frame(
  Script = c("run_rspec", "run_all_rspec", "run_slow_rspec", "run_old_rspec", "run_rspec_subset"),
  Role = c(
    "Top-level specs/*_spec.rb with spec_helper (see docs/testing.md).",
    "Compile ext/new_bridge; run specs/ and new_bridge_specs/ together.",
    "Suites under slow-specs/ (read script header for spec_helper).",
    "Legacy suites under old_specs/.",
    "Numbered subset 1–18 (Documentation/Spec_Subsets.md)."
  ),
  X2 = c("Yes", "Yes", "Yes", "Yes", "Yes"),
  stringsAsFactors = FALSE
)
bin_tbl(df_ts)
```

## Other

```{r bin-tables-other, echo=FALSE}
df_ot <- data.frame(
  Script = c("grun", "gstudio_irb.rb / gstudio_pry.rb"),
  Role = c(
    "Graal-era launcher: polyglot ruby with --jvm. Use galaaz-jruby -S instead.",
    "Loaded by gstudio; not meant to be run standalone."
  ),
  X2 = c("No", "Yes"),
  stringsAsFactors = FALSE
)
bin_tbl(df_ot)
```

For day-to-day **2.0** use, rely on **`bin/galaaz-jruby`**, **`bin/gstudio`**, **`bin/gknit`**, **`bin/run_example`**, **`bin/run_rspec`** / **`bin/run_all_rspec`**, and **`bin/galaaz-bootstrap`** on WSL when using Dockerized R. Treat **`grun`**, **`gknit_Rscript`**, and the polyglot **`ruby`** invocation in **`gknit-draft`** as **legacy** until they are ported to the same JRuby path as **`gknit`**.

# System Compatibility

Typical development and CI targets:

* **Linux** — recent Ubuntu LTS or comparable distributions (x86_64).
* **macOS** — recent releases with JRuby and GNU R available.
* **Windows** — use **WSL2** (same Linux stack as above); native Windows is not the primary target.

The native **gatekeeper** component under `ext/new_bridge` is built with `make` and a C++ toolchain; see the project `README` if compilation fails on your platform.

# Dependencies

* **JRuby** — Ruby implementation on the JVM (Galaaz 2.0 is developed and tested with JRuby).
* A **Java runtime** (JDK) compatible with your JRuby version.
* **GNU R** — `R` on your `PATH`, with ability to install packages (CRAN / Bioconductor as needed).
* **Ruby gems** — from the Galaaz repository, `bundle install` (see the `Gemfile` / gemspec).
* Optional: **Docker** — if you run R in a container (common on WSL2); see bootstrap below.
* Optional R packages for examples in this manual — e.g. `ggplot2`, `dplyr`, `knitr`, `kableExtra`, `arrow`, Bioconductor tools such as **DESeq2** (installed the usual R way).

# Installation

From a clone of the Galaaz repository:

1. Install **JRuby** and **GNU R** using your preferred package manager or Ruby version manager.
2. Install **bundler** if needed, then run **`bundle install`** in the repository root.
3. Build the bridge native code: **`make -C ext/new_bridge all`** (or **`rake compile_gatekeeper`**).
4. Run Ruby scripts with the project load path and JVM flags the project expects — the **`bin/galaaz-jruby`** wrapper sources **`bin/galaaz_jruby_env.inc.sh`** and adds **`-I lib`**. This matters especially for **Apache Arrow** integration (see `docs/testing.md`).
5. Ensure **`R`** starts GNU R and can install packages (network access to CRAN mirrors when you first call `R.install_and_loads`).

For **gKnit**, **knitr**, **rmarkdown**, and LaTeX (PDF output), install the corresponding R packages and a TeX distribution if you need PDF; the repository includes helpers such as **`bin/install-tinytex`** where appropriate.

A **table of all `bin/` scripts** (bootstrap, JRuby wrapper, gstudio, gknit, test runners, and which ones are legacy) is in the section **Command-line tools (`bin/`)** earlier in this manual.

## Windows + WSL2 (optional: Docker / R in a container)

If you run Galaaz on Windows through WSL2 and want containerized R instances,
Docker Desktop is the supported setup.

1. Install Docker Desktop on Windows:
   - https://www.docker.com/products/docker-desktop/
2. Open Docker Desktop and enable WSL integration:
   - Settings > Resources > WSL Integration
   - Enable integration for your target distro
   - Apply & Restart Docker Desktop
3. In WSL, run Galaaz bootstrap:

  > ruby bin/galaaz-bootstrap --apply
  > ruby bin/galaaz-bootstrap --check

Expected result:
- docker CLI available
- docker compose available
- docker daemon reachable (`docker info` works)

If bootstrap reports daemon is unreachable, check Docker Desktop is running and
WSL integration is enabled for the distro where Galaaz is installed.

# Usage

* Interactive shell: use 'gstudio' on the command line

  > gstudio

```{ruby simple_example}
  vec = R.c(1, 2, 3, 4)
  puts vec
```
  
* Run all specs

  > galaaz specs:all
  
* Run graphics slideshow (80+ graphics)

  > galaaz sthda:all
  
* Run labs from Introduction to Statistical Learning with R

  > galaaz islr:all

* See all available examples

  > galaaz -T
  
  Shows a list with all available executable tasks.  To execute a task, substitute the
   'rake' word in the list with 'galaaz'.  For instance, the following line shows up
  after 'galaaz -T'
  
  rake master_list:scatter_plot        # scatter_plot from:....
  
  execute
  
  > galaaz master_list:scatter_plot

# JRuby, multithreading, and the R bridge

Galaaz 2.0 runs Ruby on **JRuby**, so your application can use **real parallel threads** for
I/O-bound work (HTTP clients, database connections, message consumers, and so on). R itself is
still executed in a **single GNU R process** behind the Galaaz bridge.

When several Ruby threads call into R at the same time, the bridge **serializes** those calls:
each request is matched to a reply using an internal per-call **queue**, so you do not need to
add your own mutex around every `R.foo` from application threads. (You should still use normal
Ruby synchronization when **Ruby** data structures are shared between threads—for example, when
appending rows from each thread into a shared array before sending them to R.)

A practical pattern is:

1. Use threads (or a connection pool) to read from **multiple databases or shards** in parallel.
2. Merge the rows in Ruby under a `Mutex` if you collect into one structure.
3. Hand the merged table to R **once** (for example with `R::Arrow.from_ruby_batches` and dplyr,
   or by building a data frame) so heavy statistics run in R with fewer bridge round-trips.

A runnable sketch lives in
`examples/multithread_shards_to_r/shards_to_r.rb` (simulated shard queries; swap in your DB
driver). For concurrency tests on the bridge itself, see `specs/bridge_concurrent_spec.rb` and
`specs/arrow_from_ruby_batches_spec.rb`.

## Long-running R calls and a completion block

For R work that can take a long time, the bridge can avoid a Ruby-side **wait timeout** by
scheduling the call and resuming in a **block** when the `RET` arrives.

- **`R.eval_r_async(code, timeout: nil) { |result| ... }`** — string eval; on success, `result.value`
  is the same formatted string as **`R.eval_r`** (use `timeout: nil` for no Ruby-side limit).
- **`R::Async.<rname>(...) { |result| ... }`** — same dispatch as **`R.<rname>(...)`**, but async;
  on success, `result.value` is an **`R::Object`** (or unboxed Ruby value / Symbol), like synchronous
  **`R.<rname>`**. Optional keyword **`timeout:`** applies a Ruby-side wait limit (completion receives
  **`NewBridge::SessionClient::TimeoutError`** if R is too slow).

**Important:** **`R.foo(...) { |x| }`** is already used for dplyr-style scopes (`R::Support.new_scope`),
so async R calls must use **`R::Async`** or **`R.eval_r_async`**, not a bare **`R.foo` with a block.**

`NewBridge::EvalResult` exposes **`#ok?`**, **`#value`**, and **`#error`**. The completion block runs on a
**background thread** (not the bridge reader thread).

The example below is **plain Ruby** (no Rails). The R snippet sleeps (standing in for heavy work) and then
returns an integer so the success branch shows a **non-nil** value. (`Sys.sleep` alone returns **NULL** in R;
on success **`result.value`** is then **`nil`** in Ruby—that is expected, not a bridge error.)

```{ruby long_r_completion_block}
require 'thread'

completion = Queue.new

R.eval_r_async('({ Sys.sleep(0.3); 42L })', timeout: nil) do |result|
  if result.ok?
    puts "[completion] R finished; eval_r-style value: #{result.value.inspect}"
  else
    puts "[completion] R/bridge error: #{result.error.class}: #{result.error.message}"
  end
  completion.push(:done)
end

3.times do |i|
  puts "[main] other Ruby work step #{i + 1}"
  sleep 0.05
end

completion.pop
puts "[main] R completion has run; exiting."
```

In a **web application**, the HTTP response usually ends before R finishes, so you would not
`Queue#pop` in the controller; you would persist an identifier, let the completion block write
the outcome to storage, and notify the client (poll, WebSocket, Turbo Stream, etc.). The plain
Ruby pattern above is only to show **when** the result exists (inside the block, or after data
written there is observed elsewhere). Runnable specs live in **`new_bridge_specs/eval_r_async_spec.rb`**.

## Galaaz + Rails (JRuby) integration baseline

This section documents the baseline we used to create a working Rails app with Galaaz in WSL.
The goals were:

1. Rails boots under **JRuby**.
2. Galaaz is loaded from a local checkout (before publishing to RubyGems).
3. A request path can execute **`R.eval(...)`** and return a result.

### 1) Create the app with JRuby-friendly options

Rails defaults can pull gems that are not ideal on JRuby-first setups (for example sqlite native
extension paths and deployment extras). A minimal app avoids early friction:

```bash
cd /home/rbotafogo/desenv_linux
jruby -S rails new hedi --skip-git --minimal --skip-kamal --skip-solid --skip-active-record
```

Then install gems:

```bash
cd /home/rbotafogo/desenv_linux/hedi
jruby -S bundle install
```

### 2) Use Galaaz as a local path gem

For local development we keep a stable path:

- `~/gems/galaaz` -> symlink to your Galaaz checkout
- optional built gem archive in `~/gems/pkg/`

In Rails `Gemfile`:

```ruby
gem "galaaz", path: "/home/rbotafogo/gems/galaaz", require: false
```

And load after Rails boot in `config/application.rb`:

```ruby
config.after_initialize { require "galaaz" }
```

Why `require: false` + `after_initialize`? In this integration, loading Galaaz too early via
`Bundler.require` triggered Rails/JRuby initialization failures.

### 3) Simple request-path smoke test

A direct smoke test from Rails runner:

```bash
cd /home/rbotafogo/desenv_linux/hedi
jruby -S bundle exec rails runner "puts R.eval('sum(c(1,2,3,4,5))').inspect"
```

Expected output:

```text
15.0
```

### 4) HTTP endpoint pattern

For this baseline, a small Rack endpoint was the most stable first step to prove request-time R
evaluation. (A full ActionController stack can be enabled later as the app evolves.)

Minimal pattern:

1. Define a Rack app class under `lib/` that runs `R.eval(...)` and returns HTML/JSON.
2. Point a route to that Rack app (`root to: MyRackApp`).
3. Verify with browser/curl.

### 5) Running from WSL and opening from Windows

Recommended bind:

```bash
jruby -S bundle exec rails server -b 0.0.0.0 -p 3000
```

Then open from Windows:

- `http://localhost:3000` (usually works with WSL localhost forwarding), or
- `http://<wsl-ip>:3000` if needed.

In development, if Host Authorization blocks requests with unexpected Host headers, use:

```ruby
# config/environments/development.rb
config.hosts.clear
```

### 6) Troubleshooting checklist

- `Could not find ... in locally installed gems`:
  run `jruby -S bundle install` in the Rails app directory.
- Stale PID after crash:
  remove `tmp/pids/server.pid`.
- Local Galaaz path changed:
  verify `Gemfile` path target exists and rerun bundler.
- R runtime issues:
  confirm GNU R is installed and on `PATH` in the same shell where Rails runs.

As new Rails features are added (controllers, jobs, websockets, background rendering, plot
generation), extend this section with concrete, runnable snippets and the associated operational
checks.

# Accessing R from Ruby

One of the nice aspects of Galaaz is that variables and functions defined in R can
be easily accessed from Ruby.  For instance, to access the `mtcars` data frame from R
in Ruby, we use the symbol `:mtcars` preceded by the `~` operator: `~R[:mtcars]` retrieves the 
value of the `mtcars` object in R.

```{ruby access_r}
puts ~R[:mtcars]
```

## Scoped symbols and lexical scoping

Galaaz 2.0 uses **scoped symbols** by default. The canonical style is `R[:name]`:

- `~R[:mtcars]` fetches an R object by name.
- `R[:a] + R[:b]` builds an expression.
- `R[:year].up_to(R[:day])` builds range expressions.

If you prefer the terse `:x` syntax, you can opt in with lexical scoping using a Ruby refinement:

```ruby
module MyScript
  using Galaaz::SymbolDSL

  def self.run
    expr = :a + :b
    puts expr
    puts ~:mtcars
  end
end
```

`using Galaaz::SymbolDSL` is **lexically scoped**: only code in that module/file scope gets `:x` DSL behavior.
Outside that scope, plain Ruby `Symbol` behavior is unchanged.

To access an R function from Ruby, the R function needs to be preceded by `R.` scoping. 
Below we see an example of creating a R::Vector by calling the 'c' R function

```{ruby call_r_func}
puts vec = R.c(1.0, 2.0, 3.0, 4.0)
```
Note that 'vec' is an object of type R::Vector:

```{ruby r_object}
puts vec.class
```
Every object created by a call to an R function will be of a type that inherits from
R::Object. In R, there is also a function 'class'. In order to access that function we
can call method 'rclass' in the R::Object:

```{ruby rclass}
puts vec.rclass
```
When working with R::Object(s), it is possible to use the '.' operator to pipe operations.
When using '.', the object to which the '.' is applied becomes the first argument of the
corresponding R function. For instance, function 'c' in R, can be used to concatenate
two vectors or more vectors (in R, there are no scalar values, scalars are converted to
vectors of size 1. Within Galaaz, scalar parameter is converted to a size one vector):

```{ruby concat}
puts R.c(vec, 10, 20, 30)
```
The call above to the 'c' function can also be done using '.' notation:

```{ruby concat_with_dot}
puts vec.c(10, 20, 30)
```
We will talk about vector indexing in a later section. But notice here that indexing
an R::Vector will return another R::Vector:

```{ruby indexing}
puts vec[1]
```
Sometimes we want to index an R::Object and get back a Ruby object that is not wrapped
in an R::Object, but the native Ruby object. For this, we can index the R object with
the '>>' operator:

```{ruby native_value}
puts vec >> 0
puts vec >> 2
```

It is also possible to call an R function with named arguments, by creating the function
in Galaaz with named parameters. For instance, here is an example of creating a 'list'
with named elements:

```{ruby named_parameters}
puts R.list(first_name: "Rodrigo", last_name: "Botafogo")
```

Many R functions receive another function as argument. For instance, method 'map' applies
a function to every element of a vector. With Galaaz, it is possible to pass a Proc, 
Method or Lambda in place of the expected R function. In this next example, we will
add 2 to every element of our previously created vector:

```{ruby proc_as_param}
puts vec.map { |x| x + 2 }
```

# gKnitting a Document

This manual has been formatted using gKnit.  gKnit uses knitr and R Markdown to knit 
a document in Ruby or R and output it in any of the available formats for R Markdown.
gKnit runs with **JRuby**, **GNU R**, and Galaaz.  In gKnit, Ruby variables are persisted between 
chunks, making it an ideal solution for literate programming. Also, since it is based 
on Galaaz, Ruby chunks can have access to R variables and combining Ruby with R in one 
document is natural.

The idea of "literate programming" was first introduced by Donald Knuth in the 
1980's [@Knuth:literate_programming].
The main intention of this approach was to develop software interspersing macro snippets,
traditional source code, and a natural language such as English in a document 
that could be compiled into
executable code and at the same time easily read by a human developer. According to Knuth
"The practitioner of 
literate programming can be regarded as an essayist, whose main concern is with exposition 
and excellence of style."

The idea of literate programming evolved into the idea of reproducible research, in which
all the data, software code, documentation, graphics etc. needed to reproduce the research
and its reports could be included in a
single document or set of documents that when distributed to peers could be rerun generating
the same output and reports.

The R community has put a great deal of effort in reproducible research.  In 2002, Sweave was
introduced and it allowed mixing R code with LaTeX, generating high-quality PDF documents.  A
Sweave document could include code, the results of executing the code, graphics and text 
such that it contained the whole narrative to reproduce the research.  In
2012, Knitr, developed by Yihui Xie from RStudio was released to replace Sweave and to
consolidate in one single package the many extensions and add-on packages that
were necessary for Sweave.

With Knitr, __R markdown__ was also developed, an extension to the
Markdown format.  With __R markdown__ and Knitr it is possible to generate reports in a multitude
of formats such as HTML, Markdown, LaTeX, PDF, DVI, etc.  __R markdown__ also allows the use of
multiple programming languages such as R, Ruby, Python, etc. in the same document.  

In __R markdown__, text is interspersed with
code chunks that can be executed and both the code and its results can become
part of the final report.  Although __R markdown__ allows multiple programming languages in the
same document, only R and Python (with
the reticulate package) can persist variables between chunks.  For other languages, such as
Ruby, every chunk will start a new process and thus all data is lost between chunks, unless it
is somehow stored in a data file that is read by the next chunk.

Being able to persist data
between chunks is critical for literate programming otherwise the flow of the narrative is lost
by all the effort of having to save data and then reload it. Although this might, at first, seem like
a small nuisance, not being able to persist data between chunks is a major issue. For example, let's
take a look at the following simple example in which we want to show how to create a list and the
use it.  Let's first assume that data cannot be persisted between chunks.  In the next chunk we
create a list, then we would need to save it to file, but to save it, we need somehow to marshal the
data into a binary format:

```{ruby no_persistence}
lst = R.list(a: 1, b: 2, c: 3)
lst.saveRDS("lst.rds")
```
then, on the next chunk, where variable 'lst' is used, we need to read back it's value

```{ruby load_persisted_data}
lst = R.readRDS("lst.rds")
puts lst
```

Now, any single code has dozens of variables that we might want to use and reuse between chunks.
Clearly, such an approach becomes quickly unmanageable. Probably, because of 
this problem, it is very rare to see any __R markdown__ document in the Ruby community.

When variables can be used across chunks, then no overhead is needed:

```{ruby persistence}
lst = R.list(a: 1, b: 2, c: 3)
# any other code can be added here
```

```{ruby use_var}
puts lst
```

In the Python community, the same effort to have code and text in an integrated environment
started around the first decade of the 2000s. In 2006 IPython 0.7.2 was released.  In 2014,
Fernando Pérez spun off the Jupyter project from IPython, creating a web-based interactive
computation environment.  Jupyter can now be used with many languages, including Ruby with the
iruby gem (https://github.com/SciRuby/iruby).  In order to have multiple languages in a Jupyter
notebook the SoS kernel was developed (https://vatlab.github.io/sos-docs/).

## gKnit and __R markdown__

gKnit is based on knitr and __R markdown__ and can knit a document 
written both in Ruby and/or R and output it in any of the available formats of __R markdown__.  gKnit
allows ruby developers to do literate programming and reproducible research by allowing them to
have in a single document, text and code.

In gKnit, Ruby variables are persisted between 
chunks, making  it an ideal solution for literate programming in this language.  Also, 
since it is based on Galaaz, Ruby chunks can access R variables (`~R[:name]`, `R.*`) through the
**Galaaz bridge** while knitr drives **GNU R**—no GraalVM polyglot runtime is required.

This is not a blog post on __R markdown__, and the interested user is directed to the following links
for detailed information on its capabilities and use.

* https://rmarkdown.rstudio.com/ or
* https://bookdown.org/yihui/rmarkdown/ 

In this post, we will describe just the main aspects of __R markdown__, so the user can start 
gKnitting Ruby and R documents quickly.

## The Yaml header

An __R markdown__ document should start with a Yaml header and be stored in a file with 
'.Rmd' extension. This document has the following header for gKnitting an HTML document.

```
---
title: "How to do reproducible research in Ruby with gKnit"
author: 
    - "Rodrigo Botafogo"
    - "Daniel Mossé - University of Pittsburgh"
tags: [Tech, Data Science, Ruby, R, JRuby, Galaaz]
date: "20/02/2019"
output:
  html_document:
    self_contained: true
    keep_md: true
  pdf_document:
    includes:
      in_header: ["../../sty/galaaz.sty"]
    number_sections: yes
---
```

For more information on the options in the Yaml header, [check here](https://bookdown.org/yihui/rmarkdown/html-document.html).

## Choosing the output format when calling gknit

Yes: you can select the render target on the **command line**. **`bin/gknit`** (or **`gknit`** on your `PATH`) forwards options to **`rmarkdown::render`** via **`R::Rmarkdown.render`**.

* **`--output_format FORMAT`** — name of the format, as in the YAML `output:` block. Examples:
  * **`html_document`** — HTML (often the default you list first under `output:`).
  * **`pdf_document`** — PDF (you need a working LaTeX setup, e.g. TinyTeX; see **`bin/install-tinytex`**).
  * **`md_document`**, **`github_document`**, or any other format defined in your YAML.
  * **`all`** — render **every** format declared under `output:` in the document (same idea as in R Markdown).

If you **omit** **`--output_format`**, gknit passes **`NULL`** for the format argument. In that case **rmarkdown** uses the **first** format listed under **`output:`** in the YAML (and if none is specified there, behavior follows the usual rmarkdown defaults, typically HTML).

Other useful flags:

* **`--output_file NAME`** — output file name (optional path; see also **`--output_dir`**).
* **`--output_dir DIR`** — directory for the rendered file (created if missing).
* **`--bridge_timeout_sec`** / **`--callback_timeout_ms`** — longer R or install steps (see elsewhere in this manual).

Examples (run from the directory where paths make sense, or use absolute paths):

```text
bin/gknit blogs/manual/manual.Rmd
bin/gknit --output_format html_document blogs/manual/manual.Rmd
bin/gknit --output_format pdf_document blogs/manual/manual.Rmd
bin/gknit --output_format all blogs/manual/manual.Rmd
```

Use **`gknit -h`** for the full option list.

## __R Markdown__ formatting

Document formatting can be done with simple markups such as:

## Headers

```
# Header 1

## Header 2

### Header 3

```

## Lists

```
Unordered lists:

* Item 1
* Item 2
    + Item 2a
    + Item 2b
```

```
Ordered Lists

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```

For more R markdown formatting go to https://rmarkdown.rstudio.com/authoring_basics.html.

## R chunks

Running and executing Ruby and R code is actually what really interests us is this blog.  
Inserting a code chunk is done by adding code in a block delimited by three back ticks 
followed by an open
curly brace ('{') followed with the engine name (r, ruby, rb, include, ...), an 
any optional chunk_label and options, as shown below:

````
```{engine_name [chunk_label], [chunk_options]}`r ''`
```
````

for instance, let's add an R chunk to the document labeled 'first_r_chunk'.  This is
a very simple code just to create a variable and print it out, as follows:

````
```{r first_r_chunk}`r ''`
vec <- c(1, 2, 3)
print(vec)
```
````

If this block is added to an __R markdown__ document and gKnitted the result will be:

```{r first_r_chunk}
vec <- c(1, 2, 3)
print(vec)
```

Now let's say that we want to do some analysis in the code, but just print the result and not the
code itself.  For this, we need to add the option 'echo = FALSE'.

````
```{r second_r_chunk, echo = FALSE}`r ''`
vec2 <- c(10, 20, 30)
vec3 <- vec * vec2
print(vec3)     
```
````
Here is how this block will show up in the document. Observe that the code is not shown
and we only see the execution result in a white box

```{r second_r_chunk, echo = FALSE}
vec2 <- c(10, 20, 30)
vec3 <- vec * vec2
print(vec3)     
```

A description of the available chunk options can be found in https://yihui.name/knitr/.

Let's add another R chunk with a function definition.  In this example, a vector 
'r_vec' is created and
a new function 'reduce_sum' is defined.  The chunk specification is

````
```{r data_creation}`r ''`
r_vec <- c(1, 2, 3, 4, 5)

reduce_sum <- function(...) {
  Reduce(sum, as.list(...))
}
```
````

and this is how it will look like once executed.  From now on, to be concise in the
presentation we will not show chunk definitions any longer.


```{r data_creation}
r_vec <- c(1, 2, 3, 4, 5)

reduce_sum <- function(...) {
  Reduce(sum, as.list(...))
}
```

We can, possibly in another chunk, access the vector and call the function as follows:

```{r using_previous}
print(r_vec)
print(reduce_sum(r_vec))
```
## R Graphics with ggplot

In the following chunk, we create a bubble chart in R using ggplot and include it in 
this document.  Note that there is no directive in the code to include the image, this
occurs automatically.  The 'mpg' dataframe is natively available to R and to Galaaz as
well.

For the reader not knowledgeable of ggplot, ggplot is a graphics library based on "the
grammar of graphics" [@Wilkinson:grammar_of_graphics]. The idea of the grammar of graphics
is to build a graphics by adding layers to the plot.  More information can be found in
https://towardsdatascience.com/a-comprehensive-guide-to-the-grammar-of-graphics-for-effective-visualization-of-multi-dimensional-1f92b4ed4149.

In the plot below the 'mpg' dataset from base R is used. "The data concerns city-cycle fuel 
consumption in miles per gallon, to be predicted in terms of 3 multivalued discrete and 5 
continuous attributes." (Quinlan, 1993)

First, the 'mpg' dataset if filtered to extract only cars from the following manumactures: Audi, Ford,
Honda, and Hyundai and stored in the 'mpg_select' variable.  Then, the selected dataframe is passed
to the ggplot function specifying in the aesthetic method (aes) that 'displacement' (disp) should
be plotted in the 'x' axis and 'city mileage' should be on the 'y' axis.  In the 'labs' layer we
pass the 'title' and 'subtitle' for the plot.  To the basic plot 'g', geom\_jitter is added, that
plots cars from the same manufactures with the same color (col=manufactures) and the size of the
car point equal its high way consumption (size = hwy).  Finally, a last layer is plotter containing
a linear regression line (method = "lm") for every manufacturer.

```{r bubble, dev='png'}
# load package and data
library(ggplot2)
data(mpg, package="ggplot2")

mpg_select <- mpg[mpg$manufacturer %in% c("audi", "ford", "honda", "hyundai"), ]

# Scatterplot
theme_set(theme_bw())  # pre-set the bw theme.
g <- ggplot(mpg_select, aes(displ, cty)) + 
  labs(subtitle="mpg: Displacement vs City Mileage",
       title="Bubble chart")

g + geom_jitter(aes(col=manufacturer, size=hwy)) + 
  geom_smooth(aes(col=manufacturer), method="lm", se=F)
```

## Ruby chunks

Including a Ruby chunk is just as easy as including an R chunk in the document: just 
change the name of the engine to 'ruby'.  It is also possible to pass chunk options
to the Ruby engine; however, this version does not accept all the options that are
available to R chunks.  Future versions will add those options.

````
```{ruby first_ruby_chunk}`r ''`
```
````

In this example, the ruby chunk is called 'first_ruby_chunk'.  One important 
aspect of chunk labels is that they cannot be duplicated.  If a chunk label is 
duplicated, gKnit will stop with an error.

In the following chunk, variable 'a', 'b' and 'c' are standard Ruby variables 
and 'vec' and 'vec2' are two vectors created  by calling the 'c' method on the 
R module.

In Galaaz, the R module allows us to access R functions transparently.  The 'c' 
function in R, is a function that concatenates its arguments making a vector.

It 
should be clear that there is no requirement in gknit to call or use any R 
functions.  gKnit will knit standard Ruby code, or even general text without 
any code.

```{ruby split_data}
a = [1, 2, 3]
b = "US$ 250.000"
c = "The 'outputs' function"

vec = R.c(1, 2, 3)
vec2 = R.c(10, 20, 30)
```

In the next block, variables 'a', 'vec' and 'vec2' are used and printed.

```{ruby split2}
puts a
puts vec * vec2
```

Note that 'a' is a standard Ruby Array and 'vec' and 'vec2' are vectors that behave accordingly,
where multiplication works as expected.

## Inline Ruby code

When using a Ruby chunk, the code and the output are formatted in blocks as seen above.
This formatting is not always desired.  Sometimes, we want to have the results of the
Ruby evaluation included in the middle of a phrase. gKnit allows adding inline Ruby code 
with the 'rb' engine.  The following chunk specification will 
create and inline Ruby text:

````
This is some text with inline Ruby accessing variable 'b' which has value:
```{rb puts "```{rb puts b}\n```"}
```
and is followed by some other text!
````

<div style="margin-bottom:30px;">
</div>

This is some text with inline Ruby accessing variable 'b' which has value:
```{rb puts b}
```
and is followed by some other text!

<div style="margin-bottom:30px;">
</div>

Note that it is important not to add any new line before of after the code
block if we want everything to be in only one line, resulting in the following sentence 
with inline Ruby code.


```{ruby heading, echo = FALSE}
outputs "### #{c}"
```

He have previously used the standard 'puts' method in Ruby chunks in order produce
output.  The result of a 'puts', as seen in all previous chunks that use it,  is formatted 
inside a white box that 
follows the code block. Many times however, we would like to do some processing in the 
Ruby chunk and have the result of this processing generate and output that is 
"included" in the document as if we had typed it in __R markdown__ document.

For example, suppose we want to create a new heading in our document, but the heading 
phrase is the result of some code processing: maybe it's the first line of a file we are
going to read.  Method 'outputs' adds its output as if typed in the __R markdown__ document. 

Take now a look at variable 'c' (it was defined in a previous block above) as 
'c = "The 'outputs' function".  "The 'outputs' function" is actually the name of this
section and it was created using the 'outputs' function inside a Ruby chunk.

The ruby chunk to generate this heading is:

````
```{ruby heading}`r ''`
outputs "### #{c}"
```
````

The three '###' is the way we add a Heading 3 in __R markdown__.


### HTML Output from Ruby Chunks

We've just seen the use of method 'outputs' to add text to the the __R markdown__ 
document.  This technique can also be used to add HTML code to the document. In 
__R markdown__, any html code typed directly in the document will be properly rendered.  
Here, for instance, is a table definition in HTML and its output in the document:

```
<table style="width:100%">
  <tr>
    <th>Firstname</th>
    <th>Lastname</th> 
    <th>Age</th>
  </tr>
  <tr>
    <td>Jill</td>
    <td>Smith</td> 
    <td>50</td>
  </tr>
  <tr>
    <td>Eve</td>
    <td>Jackson</td> 
    <td>94</td>
  </tr>
</table>
```
<div style="margin-bottom:30px;">
</div>

<table style="width:100%">
  <tr>
    <th>Firstname</th>
    <th>Lastname</th> 
    <th>Age</th>
  </tr>
  <tr>
    <td>Jill</td>
    <td>Smith</td> 
    <td>50</td>
  </tr>
  <tr>
    <td>Eve</td>
    <td>Jackson</td> 
    <td>94</td>
  </tr>
</table>

<div style="margin-bottom:30px;">
</div>

But manually creating HTML output is not always easy or desirable, specially
if we intend the document to be rendered in other formats, for example, as LaTeX.
Also, The above 
table looks ugly.  The 'kableExtra' library is a great library for 
creating beautiful tables. Take a look at https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

In the next chunk, we output the 'mtcars' dataframe from R in a nicely formatted 
table.  Note that we retrieve the mtcars dataframe by using '~R[:mtcars]'.

```{ruby nice_table}
R.install_and_loads('kableExtra')
outputs (~R[:mtcars]).kable.kable_styling
```

## Including Ruby files in a chunk

R is a language that was created to be easy and fast for statisticians to use.  As far
as I know, it was not a
language to be used for developing large systems.  Of course, there are large systems and
libraries in R, but the focus of the language is for developing statistical models and
distribute that to peers.

Ruby on the other hand, is a language for large software development.  Systems written in
Ruby will have dozens, hundreds or even thousands of files.  To document a 
large system with literate programming, we cannot expect the developer to add all the 
files in a single '.Rmd' file.  gKnit provides the 'include' chunk engine to include 
a Ruby file as if it had being typed in the '.Rmd' file.

To include a file, the following chunk should be created, where <filename> is the name of
the file to be included and where the extension, if it is '.rb', does not need to be added.
If the 'relative' option is not included, then it is treated as TRUE.  When 'relative' is
true, ruby's 'require\_relative' semantics is used to load the file, when false, Ruby's 
\$LOAD_PATH is searched to find the file and it is 'require'd.

````
```{include <filename>, relative = <TRUE/FALSE>}`r ''`
```
````

Below we include file 'model.rb', which is in the same directory of this blog.  
This code uses R 'caret' package to split a dataset in a train and test sets.
The 'caret' package is a very important a useful package for doing Data Analysis,
it has hundreds of functions for all steps of the Data Analysis workflow.  To
use 'caret' just to split a dataset is like using the proverbial cannon to 
kill the fly.  We use it here only to show that integrating Ruby and R and 
using even a very complex package as 'caret' is trivial with Galaaz.

A word of advice: the 'caret' package has lots of dependencies and installing
it in a Linux system is a time consuming operation.  Method 'R.install_and_loads'
will install the package if it is not already installed and can take a while.

````
```{include model}`r ''`
```
````

```{include model}
```

```{ruby model_partition}
mtcars = ~R[:mtcars]
model = Model.new(mtcars, percent_train: 0.8)
model.partition(:mpg)
puts model.train.head
puts model.test.head
```

## Documenting Gems

gKnit also allows developers to document and load files that are not in the same directory
of the '.Rmd' file.

Here is an example of loading Ruby’s standard library file `find.rb`. In this example, relative
is set to FALSE, so Ruby will look for the file in its `$LOAD_PATH`, and the user does not
need to know its directory on disk.

````
```{include find, relative = FALSE}`r ''`
```
````

```{include find, relative = FALSE}
```

## Converting to PDF

One of the beauties of knitr is that the same input can be converted to many different outputs.
One very useful format, is, of course, PDF.  In order to converted an __R markdown__ file to PDF
it is necessary to have LaTeX installed on the system.  We will not explain here how to
install LaTeX as there are plenty of documents on the web showing how to proceed.

gKnit comes with a simple LaTeX style file for gknitting this blog as a PDF document.  Here is
the Yaml header to generate this blog in PDF format instead of HTML:

```
---
title: "gKnit - Ruby and R Knitting with Galaaz"
author: "Rodrigo Botafogo"
tags: [Galaaz, Ruby, R, JRuby, knitr, gknit]
date: "29 October 2018"
output:
  pdf\_document:
    includes:
      in\_header: ["../../sty/galaaz.sty"]
    number\_sections: yes
---
```

## Template based documents generation

When a document is converted to PDF it follows a certain conversion template. We've seen above
the use of 'galaaz.sty' as a basic template to generate a PDF document.  Using the 
'gknit-draft' app that comes with Galaaz, the same .Rmd file can be compiled to different 
looking PDF documents. Galaaz automatically loads the 'rticles' R package that comes with
templates for the following journals with the respective template name:

* ACM articles: acm_article
* ACS articles: acs_article
* AEA journal submissions: aea_article
* AGU journal submissions: ????
* AMS articles: ams_article
* American Statistical Association: asa_article
* Biometrics articles: biometrics_article
* Bulletin de l'AMQ journal submissions: amq_article
* CTeX documents: ctex
* Elsevier journal submissions: elsevier_article
* IEEE Transaction journal submissions: ieee_article
* JSS articles: jss_article
* MDPI journal submissions: mdpi_article
* Monthly Notices of the Royal Astronomical Society articles: mnras_article
* NNRAS journal submissions: nmras_article
* PeerJ articles: peerj_article
* Royal Society Open Science journal submissions: rsos_article
* Royal Statistical Society: rss_article
* Sage journal submissions: sage_article
* Springer journal submissions: springer_article
* Statistics in Medicine journal submissions: sim_article
* Copernicus Publications journal submissions: copernicus_article
* The R Journal articles: rjournal_article
* Frontiers articles: ???
* Taylor & Francis articles: ???
* Bulletin De L'AMQ: amq_article
* PLOS journal: plos_article 
* Proceedings of the National Academy of Sciences of the USA: pnas_article

In order to create a document with one of those templates, use the following command:

```
gknit-draft --filename <my_document> --template <template> --package <package>
            --create_dir
```
So, in order to create a template for writing an R Journal, use:

```
gknit-draft --filename my_r_article --template rjournal_article --package rticles
            --create_dir
```

# Accessing R variables

Galaaz allows Ruby to access variables created in R.  For example, the `mtcars` data set is 
available in R and can be accessed from Ruby by using the tilde operator followed by the
symbol for the variable, in this case `:mtcars`.  In the code below, method `outputs` is 
used to output the `mtcars` data set nicely formatted in HTML by use of the `kable` and
`kable_styling` functions. Method `outputs` is only available when used with gKnit.

```{ruby view_kable}
outputs (~R[:mtcars]).kable.kable_styling
```

# Basic Data Types

## Vector

Vectors can be thought of as contiguous cells containing data. Cells are accessed through
indexing operations such as x[5]. Galaaz has six basic (‘atomic’) vector types: logical, 
integer, real, complex, string (or character) and raw. The modes and storage modes for the 
different vector types are listed in the following
table.

| typeof    | mode      | storage.mode |
|-----------|:---------:|-------------:|
| logical   | logical   |      logical |
| integer   | numeric   |      integer |
| double    | numeric   |       double |
| complex   | complex   |      complex |
| character | character |    character |
| raw       | raw       |          raw |

Single numbers, such as 4.2, and strings, such as "four point two" are still vectors, of length
1; there are no more basic types. Vectors with length zero are possible (and useful).
String vectors have mode and storage mode "character". A single element of a character
vector is often referred to as a character string.

To create a vector the 'c' (concatenate) method from the 'R' module should be used:

```{ruby integer}
vec = R.c(1, 2, 3)
puts vec
```

Lets take a look at the type, mode and storage.mode of our vector vec.  In order to print
this out, we are creating a data frame 'df' and printing it out.  A data frame, for those
not familiar with it, is basically a table.  Here we create the data frame and add the 
column name by passing named parameters for each column, such as 'typeof:', 'mode:' and
'storage__mode?'.  You should also note here that the double underscore is converted to a '.'.
So, when printed 'storage\_\_mode' will actually print as 'storage.mode'.

Data frames will later be more carefully described.  In R, the method used to create a 
data frame is 'data.frame', in Galaaz we use 'data\_\_frame'.

```{ruby typeof_integer}
df = R.data__frame(typeof: vec.typeof, mode: vec.mode, storage__mode: vec.storage__mode)
puts df
```

If you want to create a vector with floating point numbers, then we need at least one of the
vector's element to be a float, such as 1.0.  R users should be careful, since in R a number
like '1' is converted to float and to have an integer the R developer will use '1L'. Galaaz
follows normal Ruby rules and the number 1 is an integer and 1.0 is a float.

```{ruby float}
vec = R.c(1.0, 2, 3)
puts vec
```

```{ruby typeof_float}
df = R.data__frame(typeof: vec.typeof, mode: vec.mode, storage__mode: vec.storage__mode)
outputs df.kable.kable_styling
```

In this next example we try to create a vector with a variable 'hello' that has not yet
being defined.  This will raise an exception that is printed out.  We get two return blocks,
the first with a message explaining what went wrong and the second with the full backtrace
of the error.

```{ruby r_exception}
vec = R.c(1, hello, 5)
```

Here is a vector with logical values

```{ruby logical_vector}
vec = R.c(true, true, false, false, true)
puts vec
```

### Combining Vectors

The 'c' functions used to create vectors can also be used to combine two vectors:

```{ruby combining_vectors}
vec1 = R.c(10.0, 20.0, 30.0)
vec2 = R.c(4.0, 5.0, 6.0)
vec = R.c(vec1, vec2)
puts vec
```
In galaaz, methods can be chainned (somewhat like the pipe operator in R %>%, but more generic).
In this next example, method 'c' is chainned after 'vec1'.  This also looks like 'c' is a 
method of the vector, but in reallity, this is actually closer to the pipe operator.  When
Galaaz identifies that 'c' is not a method of 'vec' it actually tries to call 'R.c' with 
'vec1' as the first argument concatenated with all the other available arguments.  The code
below is automatically converted to the code above.

```{ruby chainning_methods}
vec = vec1.c(vec2)
puts vec
```

### Vector Arithmetic

Arithmetic operations on vectors are performed element by element:

```{ruby vec_arith1}
puts vec1 + vec2
```

```{ruby mult}
puts vec1 * 5
```

When vectors have different length, a recycling rule is applied to the shorter vector:

```{ruby recycle}
vec3 = R.c(1.0, 2.0, 3.0, 4.0, 5.0, 6.0, 7.0, 8.0, 9.0)
puts vec4 = vec1 + vec3
```

### Vector Indexing

Vectors can be indexed by using the '[]' operator:

```{ruby index}
puts vec4[3]
```

We can also index a vector with another vector.  For example, in the code below, we take elements
1, 3, 5, and 7 from vec3:

```{ruby index_by_vector}
puts vec4[R.c(1, 3, 5, 7)]
```

Repeating an index and having indices out of order is valid code:

```{ruby repeated_index}
puts vec4[R.c(1, 3, 3, 1)]
```

It is also possible to index a vector with a negative number or negative vector.  In these cases
the indexed values are not returned:

```{ruby neg_index}
puts vec4[-3]
puts vec4[-R.c(1, 3, 5, 7)]
```

If an index is out of range, a missing value (NA) will be reported.

```{ruby out_of_range}
puts vec4[30]
```

It is also possible to index a vector by range:

```{ruby range}
puts vec4[(2..5)]
```

Elements in a vector can be named using the 'names' attribute of a vector: 

```{ruby naming}
full_name = R.c("Rodrigo", "A", "Botafogo")
full_name.names = R.c("First", "Middle", "Last")
puts full_name
```

Or it can also be named by using the 'c' function with named paramenters:

```{ruby named_param}
full_name = R.c(First: "Rodrigo", Middle: "A", Last: "Botafogo")
puts full_name
```

### Extracting Native Ruby Types from a Vector

Vectors created with 'R.c' are of class R::Vector.  You might have noticed that when indexing a
vector, a new vector is returned, even if this vector has one single element. In order to use
R::Vector with other ruby classes it might be necessary to extract the actual Ruby native type
from the vector. In order to do this extraction the '>>' operator is used.

```{ruby ruby_native}
puts vec4
puts vec4 >> 0
puts vec4 >> 4
```

Note that indexing with '>>' starts at 0 and not at 1, also, we cannot do negative indexing.

## Matrix

A matrix is a collection of elements organized as a two dimensional table.  A matrix can be 
created by the 'matrix' function:

```{ruby matrix}
mat = R.matrix(R.c(1.0, 2.0, 3.0, 4.0, 5.0, 6.0, 7.0, 8.0, 9.0),
               nrow: 3,
               ncol: 3)

puts mat
```
Note that matrices data is organized by column first. It is possible to organize the matrix
memory by row first passing an extra argument to the 'matrix' function:

```{ruby matrix_rowfirst}
mat_row = R.matrix(R.c(1.0, 2.0, 3.0, 4.0, 5.0, 6.0, 7.0, 8.0, 9.0),
                   nrow: 3,
                   ncol: 3,
                   byrow: true)

puts mat_row
```

### Indexing a Matrix

A matrix can be indexed by [row, column]:

```{ruby matrix_index}
puts mat_row[1, 1]
puts mat_row[2, 3]
```
It is possible to index an entire row or column with the ':all' keyword

```{ruby matrix_index_all}
puts mat_row[1, :all]
puts mat_row[:all, 2]
```

Indexing with a vector is also possible for matrices. In the following example we want
rows 1 and 3 and columns 2 and 3 building a 2 x 2 matrix.

```{ruby matrix_index_vector}
puts mat_row[R.c(1, 3), R.c(2, 3)]
```

Matrices can be combined with functions 'rbind': 

```{ruby matrix_combine_rbind}
puts mat_row.rbind(mat)
```

and 'cbind':

```{ruby matrix_combine_cbind}
puts mat_row.cbind(mat)
```

## List

A list is a data structure that can contain sublists of different types, while vector and matrix
can only hold one type of element.

```{ruby list}
nums = R.c(1.0, 2.0, 3.0)
strs = R.c("a", "b", "c", "d")
bool = R.c(true, true, false)
lst = R.list(nums: nums, strs: strs, bool: bool)
puts lst
```

Note that 'lst' elements are named elements.


### List Indexing

List indexing, also called slicing, is done using the '[]' operator and the '[[]]' operator. Let's
first start with the '[]' operator. The list above has three sublist indexing with '[]' will 
return one of the sublists.

```{ruby list_indexing}
puts lst[1]
```

Note that when using '[]' a new list is returned.  When using the double square bracket operator
the value returned is the actual element of the list in the given position and not a slice of
the original list


```{ruby list_indexing_single}
puts lst[[1]]
```

When elements are named, as dones with lst, indexing can be done by name:

```{ruby list_indexing_by_name}
puts lst[['bool']][[1]] >> 0
```

In this example, first the 'bool' element of the list was extracted, not as a list, but as a vector,
then the first element of the vector was extracted (note that vectors also accept the '[[]]' 
operator) and then the vector was indexed by its first element, extracting the native Ruby type.


## Data Frame

A data frame is a table like structure in which each column has the same number of 
rows. Data frames are the basic structure for storing data for data analysis.  We have already
seen a data frame previously when we accessed variable '~R[:mtcars]'.  In order to create a
data frame, function 'data__frame' is used:

```{ruby dataframe}
df = R.data__frame(
  year: R.c(2010, 2011, 2012),
  income: R.c(1000.0, 1500.0, 2000.0))

puts df
```

### Data Frame Indexing

A data frame can be indexed the same way as a matrix, by using '[row, column]', where row and
column can either be a numeric or the name of the row or column

```{ruby dataframe_index}
puts (~R[:mtcars]).head
puts (~R[:mtcars])[1, 2]
puts (~R[:mtcars])['Datsun 710', 'mpg']
```

Extracting a column from a data frame as a vector can be done by using the double square bracket
operator:

```{ruby dataframe_column}
puts (~R[:mtcars])[['mpg']]
```

A data frame column can also be accessed as if it were an instance variable of the data frame:

```{ruby dataframe_instance_variable}
puts (~R[:mtcars]).mpg
```

Slicing a data frame can be done by indexing it with a vector (we use 'head' to reduce the
output):

```{ruby dataframe_column_slice}
puts (~R[:mtcars])[R.c('mpg', 'hp')].head
```

A row slice can be obtained by indexing by row and using the ':all' keyword for the column:

```{ruby dataframe_row_slice}
puts (~R[:mtcars])[R.c('Datsun 710', 'Camaro Z28'), :all]
```

Finally, a data frame can also be indexed with a logical vector.  In this next example, the
'am' column of :mtcars is compared with 0 (with method 'eq').  When 'am' is equal to 0 the
car is automatic.  So, by doing '(~R[:mtcars]).am.eq 0' a logical vector is created with 
'true' whenever 'am' is 0 and 'false' otherwise.

```{ruby logical_vector_filter}
# obtain a vector with 'true' for cars with automatic transmission
automatic = (~R[:mtcars]).am.eq 0
puts automatic
```

Using this logical vector, the data frame is indexed, returning a new data frame in 
which all cars have automatic transmission.

```{ruby dataframe_logical}
# slice the data frame by using this vector
puts (~R[:mtcars])[automatic, :all]
```

# Writing Expressions in Galaaz

Galaaz extends Ruby to work with complex expressions, similar to R's expressions build with 'quote' 
(base R) or 'quo' (tidyverse).  Let's take a look at some of those expressions.

## Expressions from operators

The code below 
creates an expression summing two symbols

```{ruby expressions}
exp1 = R[:a] + R[:b]
puts exp1
```
We can build any complex mathematical expression

```{ruby expr2}
exp2 = (R[:a] + R[:b]) * 2.0 + R[:c] ** 2 / R[:z]
puts exp2
```

It is also possible to use inequality operators in building expressions

```{ruby expr3}
exp3 = (R[:a] + R[:b]) >= :z
puts exp3
```

Galaaz provides both symbolic representations for operators, such as (>, <, !=) as functional 
notation for those operators such as (.gt, .ge, etc.).  So the same expression written 
above can also be written as

```{ruby expr4}
exp4 = (R[:a] + R[:b]).ge :z
puts exp4
```

Two type of expression can only be created with the functional representation of the operators, 
those are expressions involving '==', and '='.  In order to write an expression involving '==' we
need to use the method '.eq' and for '=' we need the function '.assign'

```{ruby expr5}
exp5 = (R[:a] + R[:b]).eq :z
puts exp5
```

```{ruby expr6}
exp6 = R[:y].assign R[:a] + R[:b]
puts exp6
```
In general we think that using the functional notation is preferable to using the 
symbolic notation as otherwise, we end up writing invalid expressions such as

```{ruby exp_wrong, warning=FALSE, eval=FALSE}
exp_wrong = (R[:a] + R[:b]) == :z
puts exp_wrong
```
and it might be difficult to understand what is going on here.  The problem lies with the fact that
when using '==' we are comparing expression (R[:a] + R[:b]) to expression :z with '=='.  When the 
comparison is executed, the system tries to evaluate :a, :b and :z, and those symbols at 
this time are not bound to anything and we get a "object 'a' not found" message.
If we only use functional notation, this type of error will not occur.

## Expressions with R methods

It is often necessary to create an expression that uses a method or function.  For instance, in
mathematics, it's quite natural to write an expressin such as $y = sin(x)$. In this case, the
'sin' function is part of the expression and should not immediately executed. Now, let's say
that 'x' is an angle of 45$^\circ$ and we acttually want our expression to be $y = 0.850...$.
When we want the function to be part of the expression, we call the function preceeding it
by the letter E, such as 'E.sin(x)'

```{ruby method_expression}
exp7 = R[:y].assign E.sin(R[:x])
puts exp7
```

Expressions can also be written using '.' notation:

```{ruby expression_with_dot}
exp8 = R[:y].assign R[:x].sin
puts exp8
```

When a function has multiple arguments, the first one can be used before the '.':

```{ruby expression_multiple_args}
exp9 = R[:x].c(R[:y])
puts exp9
```

## Evaluating an Expression

Expressions can be evaluated by calling function 'eval' with a binding. A binding can be provided
with a list:

```{ruby eval_expression_list}
exp = (R[:a] + R[:b]) * 2.0 + R[:c] ** 2 / R[:z]
puts exp.eval(R.list(a: 10, b: 20, c: 30, z: 40))
```

... with a data frame:

```{ruby eval_expression_df}
df = R.data__frame(
  a: R.c(1, 2, 3),
  b: R.c(10, 20, 30),
  c: R.c(100, 200, 300),
  z: R.c(1000, 2000, 3000))

puts exp.eval(df)
```

# Manipulating Data

One of the major benefits of Galaaz is to bring strong data manipulation to Ruby. The following
examples were extracted from Hadley's "R for Data Science" (https://r4ds.had.co.nz/). This
is a highly recommended book for those not already familiar with the 'tidyverse' style of
programming in R. In the sections to follow, we will limit ourselves to convert the R code to
Galaaz. 

For these
examples, we will investigate the nycflights13 data set available on the package by the
same name.  We use function 'R.install\_and\_loads' that checks if the library is available 
locally, and if not, installs it. This data frame contains all 336,776 flights that 
departed from New York City in 2013. The data comes from the US Bureau of 
Transportation Statistics.

Dplyr often uses **tibbles** in place of classic data frames. In Galaaz, printing may differ from
the R console; if you need a classic tabular printout, convert with **`as__data__frame`** (or use
`head` / `str` in R via `R` calls).

```{ruby nycflights13}
R.install_and_loads('nycflights13')
R.library('dplyr')
```

```{ruby flights}
flights = ~R[:flights]
puts flights.head
```

## Filtering rows with Filter

In this example we filter the flights data set by giving to the filter function two expressions:
the first R[:month].eq 1

```{ruby filter_rows}
puts flights.filter((R[:month].eq 1), (R[:day].eq 1)).head
```

## Logical Operators

All flights that departed in November of December

```{ruby nov_dec}
puts flights.filter((R[:month].eq 11) | (R[:month].eq 12)).head
```

The same as above, but using the 'in' operator. In R, it is possible to define many operators
by doing %<op>%. The %in% operator checks if a value is in a vector.  In order to use those
operators from Galaaz the '._' method is used, where the first argument is the operator's
symbol, in this case ':in' and the second argument is the vector:

```{ruby in_op}
puts flights.filter(R[:month]._ :in, R.c(11, 12)).head
```

## Filtering with NA (Not Available)

Let's first create a 'tibble' with a Not Available value (R::NA).  Tibbles are a modern 
version of a data frame and operate very similarly to one.  It differs in how it outputs
the values and the result of some subsetting operations that are more consistent than
what is obtained from data frame.

```{ruby na_tibble}
df = R.tibble(x: R.c(1, R::NA, 3))
puts df
```

Now filtering by :x > 1 shows all lines that satisfy this condition, where the row with R:NA does
not.

```{ruby filter_na}
puts df.filter(R[:x] > 1)
```

To match an NA use method 'is__na'

```{ruby with_na}
puts df.filter((R[:x].is__na) | (R[:x] > 1))
```

## Arrange Rows with arrange

Arrange reorders the rows of a data frame by the given arguments.

```{ruby arrange}
puts flights.arrange(:year, :month, :day).head
```

To arrange in descending order, use function 'desc'

```{ruby desc_arrange}
puts flights.arrange(R[:dep_delay].desc).head
```

## Selecting columns

To select specific columns from a dataset we use function 'select':

```{ruby select}
puts flights.select(:year, :month, :day).head
```

It is also possible to select column in a given range

```{ruby select_range}
puts flights.select(R[:year].up_to(R[:day])).head
```

Select all columns that start with a given name sequence

```{ruby select_starts_with}
puts flights.select(E.starts_with('arr')).head
```

Other functions that can be used:

* ends_with("xyz"): matches names that end with “xyz”.

* contains("ijk"): matches names that contain “ijk”.

* matches("(.)\\1"): selects variables that match a regular expression. This one matches 
  any variables that contain repeated characters.
  
* num_range("x", (1..3)): matches x1, x2 and x3

A helper function that comes in handy when we just want to rearrange column order is 'Everything':

```{ruby everything}
puts flights.select(:year, :month, :day, E.everything).head
```

## Add variables to a dataframe with 'mutate'

```{ruby small_flights}
flights_sm = flights.
               select((R[:year].up_to(R[:day])),
                      E.ends_with('delay'),
                      :distance,
                      :air_time)

puts flights_sm.head
```

```{ruby mutate}
flights_sm = flights_sm.
               mutate(gain: R[:dep_delay] - R[:arr_delay],
                      speed: R[:distance] / R[:air_time] * 60)
puts flights_sm.head
```

## Summarising data

Function 'summarise' calculates summaries for the data frame. When no 'group_by' is used
a single value is obtained from the data frame:

```{ruby summarise}
puts flights.summarise(delay: E.mean(:dep_delay, na__rm: true))
```

When a data frame is grouped with 'group_by' summaries apply to the given group:

```{ruby summarise_group_by}
by_day = flights.group_by(:year, :month, :day)
puts by_day.summarise(delay: R[:dep_delay].mean(na__rm: true)).head
```

Next we put many operations together by pipping them one after the other:

```{ruby pipping}
delays = flights.
           group_by(:dest).
           summarise(
             count: E.n,
             dist: R[:distance].mean(na__rm: true),
             delay: R[:arr_delay].mean(na__rm: true)).
           filter(R[:count] > 20, R[:dest] != "NHL")

puts delays.head
```

# Using Data Table

The next chunk converts the **nycflights13** `flights` tibble already loaded above into a
**`data.table`**. That keeps the manual offline and avoids downloading a remote CSV during gknit
(network stalls look like bridge hangs when the transfer runs inside a single R eval).

```{ruby fread}
R.library('data.table')

flights = R.as__data__table(~R[:flights])
puts flights.dim
puts R.head(flights, 12)
```

```{ruby data_table}

data_table = R.data__table(
  ID: R.c("b","b","b","a","a","c"),
  a: (1..6),
  b: (7..12),
  c: (13..18)
)

puts data_table
puts data_table.ID
```

```{ruby subset_i}
# subset rows in i
ans = flights[(R[:origin].eq "JFK") & (R[:month].eq 6)]
puts ans.head

# Get the first two rows from flights.

ans = flights[(1..2)]
puts ans

# Sort by origin asc, then dest desc (example kept commented):
# ans = flights[E.order(:origin, -(:dest))]
# puts ans.head

```

```{ruby select_j}
# Select column(s) in j
# select arr_delay column, but return it as a vector.

ans = flights[:all, :arr_delay]
puts ans.head

# arr_delay as data.table (not plain vector).

ans = flights[:all, R[:arr_delay].list]
puts ans.head

ans = flights[:all, E.list(R[:arr_delay], R[:dep_delay])]
```

# Apache Arrow

[Apache Arrow](https://arrow.apache.org/) is a **columnar** in-memory format used heavily in R
and Python for analytics. In Galaaz, **Ruby does not hold an Arrow C++ table itself**; instead you
build ordinary Ruby structures (arrays of row hashes), and **`R::Arrow.from_ruby_batches`** creates
a real **Arrow `Table` inside GNU R**. From there you use R’s **`arrow`** and **`dplyr`** packages
as usual: **`group_by`** on the Arrow table, **`summarise`** for aggregates, then **`collect()`** to
materialize a tibble when you need in-memory R rows.

That pattern matches production use: **JRuby threads** (or sequential code) assemble many rows in
Ruby; you pay **one** bridge-heavy handoff to R; **dplyr** runs vectorised work on the Arrow table
in R.

**Prerequisites:** install R packages **`arrow`** and **`dplyr`**. Run scripts with
**`bin/galaaz-jruby`** (or the same JVM flags as in **`docs/testing.md`**) so the Arrow JNI stack is
available.

## Other `R::Arrow` helpers

The Ruby module **`R::Arrow`** (see `lib/R_interface/r_arrow.rb`) also includes:

* **`R::Arrow.table_from(df)`** — wrap an R `data.frame` / tibble as an Arrow table.
* **`R::Arrow.read_feather` / `write_feather`**, **`read_parquet`**, **`dataset(path)`** — file and
  dataset IO on paths visible to R.

## Example: many Ruby rows → Arrow in R → grouped statistics

The repository test **`slow-specs/arrow_large_pipeline_spec.rb`** builds **200k rows** in parallel
(eight threads × 25,000 rows), pushes them through **`R::Arrow.from_ruby_batches`**, then checks that
**dplyr** group summaries match a Ruby reference calculation. The same logic appears below at a
**smaller scale** so this manual can knit quickly; increase `thread_count` and `rows_per_thread`
when experimenting locally.

```{ruby arrow_pipeline_example, message=FALSE, warning=FALSE}
# Scaled-down version of slow-specs/arrow_large_pipeline_spec.rb.
unless R::Support.eval("requireNamespace('arrow', quietly=TRUE) && requireNamespace('dplyr', quietly=TRUE)") == true
  puts '(Skip: need arrow + dplyr in R; use bin/galaaz-jruby outside gKnit.)'
else
  thread_count = 4
  rows_per_thread = 500
  group_count = 5

  batches = []
  mutex = Mutex.new
  threads = []

  thread_count.times do |tid|
    threads << Thread.new do
      start = tid * rows_per_thread
      local = (start...(start + rows_per_thread)).map do |i|
        {
          id: i,
          grp: "g#{i % group_count}",
          value: (i % 17) + 1,
          weight: ((i % 5) + 1) * 0.5
        }
      end
      mutex.synchronize { batches << local }
    end
  end
  threads.each(&:join)

  tbl = R::Arrow.from_ruby_batches(batches)
  puts "R class after from_ruby_batches: #{tbl.rclass}"

  grouped = R.dplyr___group_by(tbl, :grp)
  summarised = R.dplyr___summarise(
    grouped,
    n: E.n(),
    total: E.sum(:value),
    wsum: E.sum(R[:value] * R[:weight])
  )
  out = R.dplyr___collect(summarised)

  puts 'Per-group summary (first rows):'
  puts R.as__data__frame(out).head(10)

  total_n = 0
  (1..(out.nrow >> 0)).each { |i| total_n += (out[['n']][i] >> 0) }
  puts "Sum of group counts n (should equal #{thread_count * rows_per_thread}): #{total_n}"
end
```

**What to notice:** (1) Ruby only sees **`Hash`** rows and Ruby **`Thread`** objects; (2) a single
**`from_ruby_batches`** call creates the Arrow table in R; (3) **`dplyr___group_by`** /
**`dplyr___summarise`** / **`dplyr___collect`** mirror **`dplyr::group_by`** /
**`dplyr::summarise`** / **`dplyr::collect`** on an Arrow-backed table. For a lighter test, see
**`specs/arrow_from_ruby_batches_spec.rb`**; for the full-size benchmark, run
**`bin/run_slow_rspec slow-specs/arrow_large_pipeline_spec.rb`**.

# Bioconductor and DESeq2

**Bioconductor** packages are ordinary R packages installed from the Bioconductor repositories.
Galaaz does not treat them specially: once installed in **GNU R**, you load them with
**`R.library`** like any CRAN package.

## Installing Bioconductor packages

From an R session (or `R -e '...'`), use **BiocManager** (see
[bioconductor.org](https://bioconductor.org/install/)):

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("DESeq2", "airway"))
```

The **`airway`** package ships the example **`SummarizedExperiment`** used below. **DESeq2**
pulls in several dependencies; the first install can take several minutes.

## Example: DESeq2 on the airway dataset

The script **`examples/bioconductor_deseq2_airway/deseq2_airway_galaaz.rb`** is the canonical
version in the repository. Run it from the **Galaaz repository root** with JRuby, for example:

```text
bin/galaaz-jruby examples/bioconductor_deseq2_airway/deseq2_airway_galaaz.rb
```

The workflow in Ruby mirrors a standard DESeq2 vignette:

1. **`R.library('DESeq2')`** and **`R.library('airway')`**, then **`R.data('airway')`** so the
   object exists in R’s global environment.
2. **`airway = ~R[:airway]`** pulls the experiment into a Galaaz wrapper so you can pass it to R
   functions as a Ruby value.
3. **`R.DESeqDataSet(..., design: (R[:all].til R[:cell] + R[:dex]))`** builds the **`DESeqDataSet`**. The
   **`(R[:all].til R[:cell] + R[:dex])`** form is Galaaz’s way of passing the one-sided formula
   **`~ cell + dex`** (adjust for the design you need).
4. Prefilter rows with almost no counts: **`keep = R.rowSums(R.counts(dds)) >= 10`** and
   **`dds = dds[keep, :all]`**.
5. **`dds = R.DESeq(dds)`** fits the model; **`res = R.results(dds, contrast: R.c('dex', 'trt', 'untrt'))`**
   extracts the treatment contrast (adjust **`contrast`** for your experiment).
6. Summaries use normal Ruby string interpolation on **`R.nrow`**, **`R.ncol`**, **`R.colnames`**, etc.
7. **`R.pdf(...); R.plotMA(res, ...); R.dev__off`** writes DESeq2’s MA plot (path is relative to the
   process working directory—use the repo root when running the bundled script).

Related benchmarks and warm-run notes live under **`docs/deseq2_airway_benchmark.md`** and
**`examples/bioconductor_deseq2_airway/bench_*.rb`**.

Below is the full listing (same as the file in the repository). It is **not** executed while this
manual is knitted, because **DESeq2** is heavy and may be absent on the build machine.

```{ruby deseq2_airway_full_listing, eval=FALSE}
# Canonical script: examples/bioconductor_deseq2_airway/deseq2_airway_galaaz.rb
# Run: bin/galaaz-jruby examples/.../deseq2_airway_galaaz.rb (repo root).

require 'galaaz'

R.library('DESeq2')
R.library('airway')
R.data('airway')

airway = ~R[:airway]

# Build DESeq2 dataset with one-sided formula: ~ cell + dex.
dds = R.DESeqDataSet(airway, design: (R[:all].til R[:cell] + R[:dex]))

# Prefilter genes with almost no counts.
keep = R.rowSums(R.counts(dds)) >= 10
dds = dds[keep, :all]

# Fit DE model and extract treatment effect.
dds = R.DESeq(dds)
res = R.results(dds, contrast: R.c('dex', 'trt', 'untrt'))

# Compact sanity outputs for quick verification.
puts "Samples: #{R.ncol(dds)}"
puts "Genes after prefilter: #{R.nrow(dds)}"
puts "Result rows: #{R.nrow(res)}"
puts "Result columns: #{R.colnames(res)}"
puts "Significant genes (padj < 0.05): #{R.sum(res.padj < 0.05, na__rm: true)}"

res_ordered = res[R.order(res.padj), :all]
puts R.head(R.as__data__frame(res_ordered), 10)

# Standard DESeq2 plot call written to file.
R.pdf('examples/bioconductor_deseq2_airway/plotMA_galaaz.pdf')
R.plotMA(res, ylim: R.c(-5, 5))
R.dev__off
```

If **DESeq2** and **airway** are installed, the next chunk loads the data and prints a short
preview (it does **not** run **`DESeq`** so the manual knits quickly).

```{ruby deseq2_airway_smoke, message=FALSE, warning=FALSE}
unless R::Support.eval("requireNamespace('DESeq2', quietly=TRUE) && requireNamespace('airway', quietly=TRUE)")
  puts '(Skip: install DESeq2 and airway via BiocManager in R to run the full example.)'
else
  R.library('DESeq2')
  R.library('airway')
  R.data('airway')
  airway = ~R[:airway]
  puts 'airway object (head of assay / dims via R):'
  puts "ncol(samples): #{R.ncol(airway)}"
  puts R.head(R.assay(airway), 3)
end
```

# Performance

For realistic analyses, **most wall-clock time is spent inside GNU R** (model fitting, I/O inside
R, graphics). The Galaaz **bridge** adds overhead mainly from **starting a session**, **serializing
requests**, and **wrapping results** in Ruby objects—not from reimplementing R’s numerical work.

Practical tips:

* Keep **hot loops** in R or vectorized code when possible; use Ruby for orchestration, I/O, and
  glue.
* **Reuse one process**: running many short scripts cold-starts Ruby, the JVM, and R each time;
  a long-lived process or repeated calls in one run amortize setup (see benchmarks below).
* **Batch data**: merge shards in Ruby, then call **`R::Arrow.from_ruby_batches`** (or build one
  data frame) instead of millions of tiny R calls.

For measured discussion (including DESeq2-style workloads and warm comparisons), see
**`docs/performance.md`** and **`docs/deseq2_airway_benchmark.md`** in the Galaaz repository.

# Graphics in Galaaz

Creating graphics in Galaaz is quite easy, as it can use all the power of ggplot2.  There are
many resources on the web that teach ggplot, so here we give a quick example of ggplot 
integration with Ruby.  We continue to use the :mtcars dataset and we will plot a diverging
bar plot, showing cars that have 'above' or 'below' gas consumption. Let's first prepare
the data frame with the necessary data:

```{ruby diverging_plot_pre}
# :mtcars -> Ruby handle
mtcars = ~R[:mtcars]

# Row labels are not a plot column; copy them to car_name.
mtcars.car_name = R.rownames(R[:mtcars])

# Z-score mpg (mean/sd on mtcars.mpg); round to 2 decimals.
mtcars.mpg_z = ((mtcars.mpg - mtcars.mpg.mean)/mtcars.mpg.sd).round 2

# ifelse is vectorized: below / above average mpg_z.
mtcars.mpg_type = (mtcars.mpg_z < 0).ifelse("below", "above")

# Sort rows by mpg_z.
mtcars = mtcars[mtcars.mpg_z.order, :all]

# Factor car_name so plot order follows sort.
mtcars.car_name = mtcars.car_name.factor levels: mtcars.car_name

puts mtcars.head
```
Now, let's plot the diverging bar plot.  When using gKnit, you normally do **not** need to open a
graphics device manually; gKnit arranges the figure device for chunk output. Galaaz 
provides integration with ggplot. The interested reader should check online for more
information on ggplot, since it is outside the scope of this manual describing 
how ggplot works. Here we give only a brief description of how this plot is generated.

ggplot implements the 'grammar of graphics'. In this approach, plots are built by
adding layers to the plot.  On the first layer we describe what we want on the 'x'
and 'y' axis of the plot.  In this case, we have 'car_name' on the 'x' axis and 
'mpg\_z' on the 'y' axis. Then the type of graph is specified by adding
'geom\_bar' (for a bar graph).  We specify that our bars should be filled using 
'mpg\_type', which is either 'above' or 'below' giving then two colours for
filling. On the next layer we specify the labels for the graph, then we add the
title and subtitle.  Finally, in a bar chart usually bars go on the vertical direction,
but in this graph we want the bars to be horizontally laid so we add 'coord\_flip'.

```{ruby diverging_bar, fig.width = 9.1, fig.height = 6.5}
require 'ggplot'

puts mtcars.ggplot(E.aes(x: :car_name, y: :mpg_z, label: :mpg_z)) +
     R.geom_bar(E.aes(fill: :mpg_type), stat: 'identity', width: 0.5) +
     R.scale_fill_manual(name: 'Mileage',
                         labels: R.c('Above Average', 'Below Average'),
                         values: R.c('above': '#00ba38', 'below': '#f8766d')) +
     R.labs(subtitle: "Normalised mileage from 'mtcars'",
            title: "Diverging Bars") + 
     R.coord_flip
```

# Coding with Tidyverse

In R, and when coding with 'tidyverse', arguments to a function are usually not 
*referentially transparent*. That is, you can’t replace a value with a seemingly equivalent 
object that you’ve defined elsewhere. To see the problem, let's first define a data frame:

```{ruby df}
df = R.data__frame(x: (1..3), y: (3..1))
puts df
```

and now, let's look at this code:

```{r not_transp, eval=FALSE}
my_var <- x
filter(df, my_var == 1)
```
It generates the following error: "object 'x' not found.

However, in Galaaz, arguments are referentially transparent as can be seen by the 
code below.  Note initially that 'my_var = R[:x]' will not give the error "object 'x' not found" 
since ':x' is treated as an expression and assigned to my\_var. Then when doing (my\_var.eq 1), 
my\_var is a variable that resolves to ':x' and it becomes equivalent to (R[:x].eq 1) which is
what we want.

```{ruby my_var}
my_var = R[:x]
puts df.filter(my_var.eq 1)
```
As stated by Hadley

> dplyr code is ambiguous. Depending on what variables are defined where, 
> filter(df, x == y) could be equivalent to any of:

```
df[df$x == df$y, ]
df[df$x == y, ]
df[x == df$y, ]
df[x == y, ]
```
In galaaz this ambiguity does not exist, filter(df, x.eq y) is not a valid expression as 
expressions are build with symbols.  In doing filter(df, R[:x].eq y) we are looking for elements
of the 'x' column that are equal to a previously defined y variable.  Finally in 
filter(df, R[:x].eq R[:y]) we are looking for elements in which the 'x' column value is equal to
the 'y' column value. This can be seen in the following two chunks of code:

```{ruby disamb1}
y = 1
x = 2

# looking for values where the 'x' column is equal to the 'y' column
puts df.filter(R[:x].eq R[:y])
```

```{ruby disamb2}
# looking for values where the 'x' column is equal to the 'y' variable
# in this case, the number 1
puts df.filter(R[:x].eq y)
```
## Writing a function that applies to different data sets

Let's suppose that we want to write a function that receives as the first argument a data frame
and as second argument an expression that adds a column to the data frame that is equal to the
sum of elements in column 'a' plus 'x'. 

Here is the intended behaviour using the 'mutate' function of 'dplyr':

```
mutate(df1, y = a + x)
mutate(df2, y = a + x)
mutate(df3, y = a + x)
mutate(df4, y = a + x)
```
The naive approach to writing an R function to solve this problem is:

```
mutate_y <- function(df) {
  mutate(df, y = a + x)
}
```
Unfortunately, in R, this function can fail silently if one of the variables isn’t present 
in the data frame, but is present in the global environment.  We will not go through here how
to solve this problem in R.

In Galaaz the method mutate_y below will work fine and will never fail silently.

```{ruby mutate_y, warning=FALSE}
def mutate_y(df)
  # Mutate column names are Ruby kwargs (y: …). Use .assign only for R `<-` expressions.
  df.mutate(y: R[:a] + R[:x])
end
```
Here we create a data frame that has only one column named 'x':

```{ruby data_frame_no_a_column, warning=FALSE}
df1 = R.data__frame(x: (1..3))
puts df1
```

Note that method mutate_y will fail independently from the fact that variable 'a' is defined and
in the scope of the method.  Variable 'a' has no relationship with the symbol `R[:a]` used in the
definition of 'mutate\_y' above:

```{ruby call_mutate_y, warning = FALSE}
a = 10
mutate_y(df1)
```
## Different expressions

Let's move to the next problem as presented by Hadley where trying to write a function in R 
that will receive two argumens, the first a variable and the second an expression is not trivial.
Below we create a data frame and we want to write a function that groups data by a variable and
summarises it by an expression:

```{r diff_expr}
library(dplyr)
set.seed(123)

df <- data.frame(
  g1 = c(1, 1, 2, 2, 2),
  g2 = c(1, 2, 1, 2, 1),
  a = sample(5),
  b = sample(5)
)

as.data.frame(df) 

d2 <- df %>%
  group_by(g1) %>%
  summarise(a = mean(a))
	       
as.data.frame(d2)	       

d2 <- df %>%
  group_by(g2) %>%
  summarise(a = mean(a))
	       
as.data.frame(d2)	       
```

As shown by Hadley, one might expect this function to do the trick:

```{r diff_exp_fnc}
my_summarise <- function(df, group_var) {
  df %>%
    group_by(group_var) %>%
    summarise(a = mean(a))
}

# my_summarise(df, g1)
#> Error: Column `group_var` is unknown
```

In order to solve this problem, coding with dplyr requires the introduction of many new concepts
and functions such as 'quo', 'quos', 'enquo', 'enquos', '!!' (bang bang), '!!!' (triple bang). 
Again, we'll leave to Hadley the explanation on how to use all those functions.

Now, let's try to implement the same function in galaaz.  The next code block first prints the
'df' data frame defined previously in R (to access an R variable from Galaaz, we use the tilde
operator `~` applied to the R variable name as a symbol, e.g. `:df`).

```{ruby r_dataframe}
puts ~R[:df]
```

We then create the 'my_summarize' method and call it passing the R data frame and 
the group by variable 'R[:g1]':

```{ruby diff_exp_ruby_func}
def my_summarize(df, group_var)
  df.group_by(group_var).
    summarize(a: R[:a].mean)
end

puts my_summarize(~R[:df], R[:g1])
```

It works!!! Well, let's make sure this was not just some coincidence

```{ruby group_g2}
puts my_summarize(~R[:df], R[:g2])
```

Great, everything is fine! No magic, no new functions, no complexities, just normal, standard Ruby
code.  If you've ever done NSE in R, this certainly feels much safer and easy to implement.

## Different input variables

In the previous section we've managed to get rid of all NSE formulation for a simple example, but
does this remain true for more complex examples, or will the Galaaz way prove inpractical for
more complex code?

In the next example Hadley proposes us to write a function that given an expression such as 'a'
or 'a * b', calculates three summaries.  What we want a function that does the same as these R
statements:

```
summarise(df, mean = mean(a), sum = sum(a), n = n())
#> # A tibble: 1 x 3
#>    mean   sum     n
#>   <dbl> <int> <int>
#> 1     3    15     5

summarise(df, mean = mean(a * b), sum = sum(a * b), n = n())
#> # A tibble: 1 x 3
#>    mean   sum     n
#>   <dbl> <int> <int>
#> 1   9    45     5
```

Let's try it in galaaz:

```{ruby summarize_method}
def my_summarise2(df, expr)
  df.summarize(
    mean: E.mean(expr),
    sum: E.sum(expr),
    n: E.n
  )
end

puts my_summarise2((~R[:df]), :a)
puts "\n"
puts my_summarise2((~R[:df]), R[:a] * R[:b])
```

Once again, there is no need to use any special theory or functions.  The only point to be 
careful about is the use of 'E' to build expressions from functions 'mean', 'sum' and 'n'.

## Different input and output variable

Now the next challenge presented by Hadley is to vary the name of the output variables based on 
the received expression.  So, if the input expression is 'a', we want our data frame columns to
be named 'mean\_a' and 'sum\_a'.  Now, if the input expression is 'b', columns
should be named 'mean\_b' and 'sum\_b'.

```
mutate(df, mean_a = mean(a), sum_a = sum(a))
#> # A tibble: 5 x 6
#>      g1    g2     a     b mean_a sum_a
#>   <dbl> <dbl> <int> <int>  <dbl> <int>
#> 1     1     1     1     3      3    15
#> 2     1     2     4     2      3    15
#> 3     2     1     2     1      3    15
#> 4     2     2     5     4      3    15
#> # … with 1 more row

mutate(df, mean_b = mean(b), sum_b = sum(b))
#> # A tibble: 5 x 6
#>      g1    g2     a     b mean_b sum_b
#>   <dbl> <dbl> <int> <int>  <dbl> <int>
#> 1     1     1     1     3      3    15
#> 2     1     2     4     2      3    15
#> 3     2     1     2     1      3    15
#> 4     2     2     5     4      3    15
#> # … with 1 more row
```
In order to solve this problem in R, Hadley needs to introduce some more new functions and notations:
'quo_name' and the ':=' operator from package 'rlang'

Here is our Ruby code:

```{ruby name_change}
def my_mutate(df, expr)
  mean_name = "mean_#{expr.to_s}"
  sum_name = "sum_#{expr.to_s}"

  df.mutate(mean_name => E.mean(expr),
            sum_name => E.sum(expr))
end

puts my_mutate((~R[:df]), :a)
puts "\n"
puts my_mutate((~R[:df]), :b)
```
It really seems that "Non Standard Evaluation" is actually quite standard in Galaaz! But, you 
might have noticed a small change in the way the arguments to the mutate method were called.
In a previous example we used df.summarise(mean: E.mean(:a), ...) where the column name was
followed by a ':' colom.  In this example, we have df.mutate(mean_name => E.mean(expr), ...)
and variable mean\_name is not followed by ':' but by '=>'.  This is standard Ruby notation.

[explain....]

## Capturing multiple variables

Moving on with new complexities, Hadley proposes us to solve the problem in which the 
summarise function will receive any number of grouping variables.

This again is quite standard Ruby.  In order to receive an undefined number of paramenters
the paramenter is preceded by '*':

```{ruby multiple_vars}
def my_summarise3(df, *group_vars)
  df.group_by(*group_vars).
    summarise(a: E.mean(:a))
end

puts my_summarise3((~R[:df]), R[:g1], R[:g2])
```

## Why does R require NSE and Galaaz does not?

NSE introduces a number of new concepts, such as 'quoting', 'quasiquotation', 'unquoting' and 
'unquote-splicing', while in Galaaz none of those concepts are needed. What gives?

R is an extremely flexible language and it has lazy evaluation of parameters. When in R a 
function is called as 'summarise(df, a = b)', the summarise function receives the litteral
'a = b' parameter and can work with this as if it were a string. In R, it is not clear what
a and b are, they can be expressions or they can be variables, it is up to the function to
decide what 'a = b' means.

In Ruby, there is no lazy evaluation of parameters and 'a' is always a variable and so is 'b'.
Variables assume their value as soon as they are used, so 'x = a' is immediately evaluate and 
variable 'x' will receive the value of variable 'a' as soon as the Ruby statement is executed. 
Ruby also provides the notion of a symbol; ':a' is a symbol and does not evaluate to anything.
Galaaz uses Ruby symbols to build expressions that are not bound to anything: 'R[:a].eq R[:b]' is
clearly an expression and has no relationship whatsoever with the statment 'a = b'. By using
symbols, variables and expressions all the possible ambiguities that are found in R are 
eliminated in Galaaz.

The main problem that remains, is that in R, functions are not clearly documented as what type
of input they are expecting, they might be expecting regular variables or they might be 
expecting expressions and the R function will know how to deal with an input of the form 
'a = b', now for the Ruby developer it might not be immediately clear if it should call the 
function passing the value 'true' if variable 'a' is equal to variable 'b' or if it should
call the function passing the expression 'R[:a].eq R[:b]'.


## Advanced dplyr features

In the blog: Programming with dplyr by using dplyr (https://www.r-bloggers.com/programming-with-dplyr-by-using-dplyr/) Iñaki Úcar shows surprise that some R users are trying to code in dplyr avoiding
the use of NSE.  For instance he says:

> Take the example of seplyr. It stands for standard evaluation dplyr, and enables us to 
> program over dplyr without having “to bring in (or study) any deep-theory or 
> heavy-weight tools such as rlang/tidyeval”.

For me, there isn't really any surprise that users are trying to avoid dplyr deep-theory. R
users frequently are not programmers and learning to code is already hard business, on top
of that, having to learn how to 'quote' or 'enquo' or 'quos' or 'enquos' is not necessarily
a 'piece of cake'. So much so, that 'tidyeval' has some more advanced functions that instead
of using quoted expressions, uses strings as arguments.

In the following examples, we show the use of functions 'group\_by\_at', 'summarise\_at' and
'rename\_at' that receive strings as argument. The data frame used in 'starwars' that describes
features of characters in the Starwars movies:

```{ruby starwars}
puts (~R[:starwars]).head
```
The grouped_mean function below will receive a grouping variable and calculate summaries for
the value\_variables given:

```{r grouped_mean}
library(dplyr)
grouped_mean <- function(data, grouping_variables, value_variables) {
  data %>%
    group_by_at(grouping_variables) %>%
    mutate(count = n()) %>%
    summarise_at(c(value_variables, "count"), mean, na.rm = TRUE) %>%
    rename_at(value_variables, funs(paste0("mean_", .)))
    }

gm = starwars %>% 
   grouped_mean("eye_color", c("mass", "birth_year"))

as.data.frame(gm)   
```

The same code with Galaaz, becomes:

```{ruby advanced_starwars}
def grouped_mean(data, grouping_variables, value_variables)
  data.
    group_by_at(grouping_variables).
    mutate(count: E.n).
    summarise_at(E.c(value_variables, "count"), ~R[:mean], na__rm: true).
    rename_at(value_variables, E.funs(E.paste0("mean_", value_variables)))
end

puts grouped_mean((~R[:starwars]), "eye_color", E.c("mass", "birth_year"))
```

The examples above cover programmatic dplyr with string column names and `_at` helpers. The same
Galaaz patterns (symbols, `E.*` for expression-safe functions, and Ruby methods on R-backed objects)
extend to other tidyverse workflows; consult R package documentation for function-specific
arguments.

# Contributing

* Fork it
* Create your feature branch (`git checkout -b my-new-feature`)
* Write tests — use **`bin/run_rspec`** or **`bin/run_all_rspec`** with **JRuby** so JVM flags and
  the load path match **`docs/testing.md`**
* Commit your changes (`git commit -am 'Add some feature'`)
* Push to the branch (`git push origin my-new-feature`)
* Open a pull request

# References
