---
title: "How to do reproducible research in Ruby with gKnit"
author:
    - "Rodrigo Botafogo"
    - "Daniel Mossé - University of Pittsburgh"
tags: [Tech, Data Science, Ruby, R, GraalVM]
date: "20/02/2019"
output:
  pdf_document:
    includes:
      in_header: ["../../sty/galaaz.sty"]
    number_sections: yes
  html_document:
    self_contained: true
    keep_md: true
---

```{r setup, echo=FALSE}

```

# Introduction

The idea of "literate programming" was first introduced by Donald Knuth in the 1980's.
The main intention of this approach was to develop software interspersing macro snippets,
traditional source code, and a natural language such as English in a document 
that could be compiled into
executable code and at the same time easily read by a human developer. According to Knuth
"The practitioner of 
literate programming can be regarded as an essayist, whose main concern is with exposition 
and excellence of style."

The idea of literate programming evolved into the idea of reproducible research, in which
all the data, software code, documentation, graphics etc. needed to reproduce the research
and its reports could be included in a
single document or set of documents that when distributed to peers could be rerun generating
the same output and reports.

The R community has put a great deal of effort in reproducible research.  In 2002, Sweave was
introduced and it allowed mixing R code with Latex generating high quality PDF documents.  Those
documents could include the code, the result of executing the code, graphics and text.  This
contained the whole narrative to reproduce the research.  But Sweave had many problems and in
2012, Knitr, developed by Yihui Xie from RStudio was released, solving many of the long lasting
problems from Sweave and including in one single package many extensions and add-on packages that
were necessary for Sweave.

With Knitr, R markdown was also developed, an extension to the
Markdown format.  With R markdown and Knitr it is possible to generate reports in a multitude
of formats such as HTML, markdown, Latex, PDF, dvi, etc.  R markdown also allows the use of
multiple programming languages in the same document.  In R markdown text is interspersed with
code chunks that can be executed and both the code and its results can become
part of the final report.  Although R markdown allows multiple programming languages in the
same document, only R and Python (with
the reticulate package) can persist variables between chunks.  For other languages, such as
Ruby, every chunk will start a new process and thus all data is lost between chunks, unless it
is somehow stored in a data file that is read by the next chunk.

Being able to persist data
between chunks is critical for literate programming otherwise the flow of the narrative is lost
by all the effort of having to save data and then reload it. Probably, because of 
this impossibility,
it is very rare to see any R markdown document in the Ruby community.  Also, the use of 
R markdown for the Ruby community would also require the Ruby developer to download R and 
have some minimal knowledge of Knitr.

In the Python community, the same effort to have code and text in an integrated environment
started around the first decade of 2000. In 2006 iPython 0.7.2 was released.  In 2014,
Fernando Pérez, spun off project Jupyter from iPython creating a web-based interactive
computation environment.  Jupyter can now be used with many languages, including Ruby with the
iruby gem (https://github.com/SciRuby/iruby).  I am not sure if multiple languages can be used
in a Jupyter notebook and if variables can persist between chunks.

# gKnitting a Document

This document describes gKnit.  gKnit uses Knitr and R markdown to knit a document in Ruby or R
and output it in any of the available formats for R markdown.  
gKnit runs atop of GraalVM, and Galaaz (an integration 
library between Ruby and R).  In gKnit, Ruby variables are persisted between chunks, making 
it an ideal solution for literate programming in this language.  Also, since it is based on 
Galaaz, Ruby chunks can have access to R variables and Polyglot Programming with Ruby and R
is quite natural.

Galaaz has been describe already in the following posts:

* https://towardsdatascience.com/ruby-plotting-with-galaaz-an-example-of-tightly-coupling-ruby-and-r-in-graalvm-520b69e21021.  
* https://medium.freecodecamp.org/how-to-make-beautiful-ruby-plots-with-galaaz-320848058857

This is not a blog post on R markdown, and the interested user is directed to the following links
for detailed information on its capabilities and use.

* https://rmarkdown.rstudio.com/ or
* https://bookdown.org/yihui/rmarkdown/ 

Here, we will describe quickly the main aspects of R markdown, so the user can start gKnitting
Ruby and R documents quickly.

## The Yaml header

An R markdown document should start with a Yaml header and be stored in a file with 
'.Rmd' extension. This document has the following header for gKitting an HTML document.

```
---
title: "How to do reproducible research in Ruby with gKnit"
author: 
    - "Rodrigo Botafogo"
    - "Daniel Mossé - University of Pittsburgh"
tags: [Tech, Data Science, Ruby, R, GraalVM]
date: "20/02/2019"
output:
  html_document:
    self_contained: true
    keep_md: true
  pdf_document:
    includes:
      in_header: ["../../sty/galaaz.sty"]
    number_sections: yes
---
```

For more information on the options in the Yaml header, check https://bookdown.org/yihui/rmarkdown/html-document.html.

## R Markdown formatting

Document formatting can be done with simple markups such as:

### Headers

```
# Header 1

## Header 2

### Header 3

```

### Lists

```
Unordered lists:

* Item 1
* Item 2
    + Item 2a
    + Item 2b
```

```
Ordered Lists

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```

Please, go to https://rmarkdown.rstudio.com/authoring_basics.html, for more R markdown formatting.

### R chunks

Running and executing Ruby and R code is actually what really interests us is this blog.  
Inserting a code chunk is done by adding code in a block delimited by three back ticks 
followed by an open
curly brace ('{') followed with the engine name (r, ruby, rb, include, ...), an 
any optional chunk_label and options, as shown bellow:

````
```{engine_name [chunk_label], [chunk_options]}`r ''`
```
````

for instance, let's add an R chunk to the document labeled 'first_r_chunk'.  This is
a very simple code just to create a variable and print it out.  The code block should
be defined as follows:

````
```{r first_r_chunk}`r ''`
vec <- c(1, 2, 3)
print(vec)
```
````

If this block is added to an R markdown document and gKnitted the result will be:

```{r first_r_chunk}
vec <- c(1, 2, 3)
print(vec)
```

Now let's say that we want to do some analysis in the code, but just print the result and not the
code itself.  For this, we need to add the option 'echo = FALSE'.

````
```{r second_r_chunk, echo = FALSE}`r ''`
vec2 <- c(10, 20, 30)
vec3 <- vec * vec2
print(vec3)     
```
````
Here is how this block will show up in the document. Observe that the code is not shown
and we only see the execution result in a white box

```{r second_r_chunk, echo = FALSE}
vec2 <- c(10, 20, 30)
vec3 <- vec * vec2
print(vec3)     
```

A description of the available chunk options can be found in the documentation cited above.

Let's add another R chunkd with a function definition.  In this example, a vector 
'r_vec' is created and
a new function 'reduce_sum' is defined.  The chunk specification is

````
```{r data_creation}`r ''`
r_vec <- c(1, 2, 3, 4, 5)

reduce_sum <- function(...) {
  Reduce(sum, as.list(...))
}
```
````

and this is how it will look like once executed.  From now on, we will not 
show the chunk definition any longer.


```{r data_creation}
r_vec <- c(1, 2, 3, 4, 5)

reduce_sum <- function(...) {
  Reduce(sum, as.list(...))
}
```

We can, possibly in another chunk, access the vector and call the function as follows:

```{r using_previous}
print(r_vec)
print(reduce_sum(r_vec))
```
### R Graphics with ggplot

In the following chunk, we create a bubble chart in R using ggplot and include it in 
this document.  Note that there is no directive in the code to include the image, this
occurs automatically.  The 'mpg' dataframe is natively available to R and to Galaaz as
well.

```{r bubble, dev='png'}
# load package and data
library(ggplot2)
data(mpg, package="ggplot2")

mpg_select <- mpg[mpg$manufacturer %in% c("audi", "ford", "honda", "hyundai"), ]

# Scatterplot
theme_set(theme_bw())  # pre-set the bw theme.
g <- ggplot(mpg_select, aes(displ, cty)) + 
  labs(subtitle="mpg: Displacement vs City Mileage",
       title="Bubble chart")

g <- g + geom_jitter(aes(col=manufacturer, size=hwy)) + 
     geom_smooth(aes(col=manufacturer), method="lm", se=F)

```

### Ruby chunks


Including a Ruby chunk is just as easy as including an R chunk in the document: just 
change the name of the engine to 'ruby'.  It is also possible to pass chunk options
to the Ruby engine; however, this version does not accept all the options that are
available to R chunks.  Future versions will add those options.

````
```{ruby first_ruby_chunk}`r ''`
```
````

In this example, the ruby chunk is called 'first_ruby_chunk'.  One important 
aspect of chunk labels is that they cannot be duplicated.  If a chunk label is 
duplicated, gKnitting will stop with an error.

Another relevant point with Ruby chunks is that they are evaluated in the scope
of a class called RubyChunk.  To make sure that variables are
available between chunks, they should be made as instance variables of the 
RubyChunk class.  In the following chunk, variable '\@a', '\@b' and '\@c' 
are standard Ruby variables and '\@vec' and '\@vec2' are two vectors created 
by calling the 'c' method on the R module.

In Galalaaz, the R module allows us to access R functions transparently.  It 
should be clear that there is no requirement in gknit to call or use any R 
functions.  gKnit will knit standard Ruby code, or even general text without 
any code.

```{ruby split_data}
@a = [1, 2, 3]
@b = "US$ 250.000"
@c = "The 'outputs' function"

@vec = R.c(1, 2, 3)
@vec2 = R.c(10, 20, 30)
```

In this next block, variables '\@a', '\@vec' and '\@vec2' are used and printed.

```{ruby split2}
puts @a
puts @vec * @vec2
```

Note that @a is a standard Ruby Array and @vec and @vec2 are vectors that behave accordingly,
where multiplication works as expected.


### Accessing R from Ruby

One of the nice aspects of Galaaz on GraalVM, is that variables and functions defined in R, can
be easily accessed from Ruby.  This next chunk, reads data from R and uses the 'reduce_sum'
function defined previously.  To access an R variable from Ruby the '~' function should be
applied to the Ruby symbol representing the R variable.  Since the R variable is called 'r_vec',
in Ruby, the symbol to acess it is ':r_vec' and thus '~:r_vec' retrieves the value of the
variable.

```{ruby access_r}
puts ~:r_vec
```

In order to call an R function, the 'R.' module is used as follows

```{ruby call_r_func}
puts R.reduce_sum(~:r_vec)
```

### Ruby Plotting

We have seen an example of plotting with R.  Plotting with Ruby does not require
anything different from plotting with R:

```{ruby diverging_bar, fig.width = 9.1, fig.height = 6.5}
require 'ggplot'

mtcars = ~:mtcars

mtcars.car_name = mtcars.rownames  # create new column for car names
mtcars.mpg_z = ((mtcars.mpg - mtcars.mpg.mean) / mtcars.mpg.sd).round 2
mtcars.mpg_type = (mtcars.mpg_z < 0).ifelse('below', 'above')
mtcars = mtcars[mtcars.mpg_z.order, :all]
mtcars.car_name = R.factor(mtcars.car_name, levels: mtcars.car_name)

puts mtcars.ggplot(E.aes(x: :car_name, y: :mpg_z, label: :mpg_z)) +
     R.geom_bar(E.aes(fill: :mpg_type), stat: 'identity', width: 0.5) +
     R.scale_fill_manual(name: 'Mileage',
                         labels: R.c('Above Average', 'Below Average'),
                         values: R.c('above': '#00ba38', 'below': '#f8766d')) +
     R.labs(subtitle: "Normalised mileage from 'mtcars'",
            title: "Diverging Bars") + 
     R.coord_flip
```

### Inline Ruby code

When using a Ruby chunk, the code and the output are formated in blocks as seen above.
This formatting is not always desired.  Sometimes, one wants to have the result of the
Ruby evalutaion included in the middle of a phrase. gKnit allows adding inline Ruby code 
with the 'rb' engine.  The following text will 
create and inline Ruby text:

````
This is some text with inline Ruby accessing variable \@b which has value:
```{rb puts "```{rb puts @b}\n```"}
```
and is followed by some other text!
````

Note that it is important not to add any new line before of after the code
block if we want everything to be in only one line, resulting in the following sentence 
with inline Ruby code

<div style="margin-bottom:30px;">
</div>

This is some text with inline Ruby accessing variable \@b which has value:
```{rb puts @b}
```
and is followed by some other text!

<div style="margin-bottom:30px;">
</div>


```{ruby heading, echo = FALSE}
outputs "### #{@c}"
```

He have previously used the standard 'puts' method in Ruby chunks in order to get some
output.  As can be seen, the result of a 'puts' is formatted inside a white box that 
follows the code block. Many times however, we would like to do some processing in the 
Ruby chunk and have the result of this processing generate and output that is 
'included' in the document as if we had typed it in R markdown.  

For example, suppose we want to create a new 'heading' in our document, but the heading 
phrase is the result of some code processing: maybe it's the first line of a file we are
going to read.  Method 'outputs' adds its output as if typed in the R markdown document. 

Take now a look at variable '@c' (it was defined in a previous block above) as 
'@c = "The 'outputs' function".  "The 'outputs' function" is actually the name of this
section and it was created using the 'outputs' function inside a Ruby chunk.

The ruby chunk to generate this heading is:

````
```{ruby heading}`r ''`
outputs "### #{@c}"
```
````

The three '###' are the way we add a Heading 3 in R markdown.


### HTML Output from Ruby Chunks

We've just seen the use of method 'outputs' to add text to the the R markdown 
document.  This technique can also be used to add HTML code to the document. In R 
markdown any html code typed directly in the document will be properly rendered.  
Here, for instance, is a table definition in HTML and its output in the document:

```
<table style="width:100%">
  <tr>
    <th>Firstname</th>
    <th>Lastname</th> 
    <th>Age</th>
  </tr>
  <tr>
    <td>Jill</td>
    <td>Smith</td> 
    <td>50</td>
  </tr>
  <tr>
    <td>Eve</td>
    <td>Jackson</td> 
    <td>94</td>
  </tr>
</table>
```
<div style="margin-bottom:30px;">
</div>

<table style="width:100%">
  <tr>
    <th>Firstname</th>
    <th>Lastname</th> 
    <th>Age</th>
  </tr>
  <tr>
    <td>Jill</td>
    <td>Smith</td> 
    <td>50</td>
  </tr>
  <tr>
    <td>Eve</td>
    <td>Jackson</td> 
    <td>94</td>
  </tr>
</table>

<div style="margin-bottom:30px;">
</div>

But manually creating HTML output is not always easy or desirable. The above 
table certainly looks ugly.  The 'kableExtra' library is a great library for 
creating beautiful tables. Take a look at https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

In the next chunk, we output the 'mtcars' dataframe from R in a nicely formatted 
table.  Note that we retrieve the mtcars dataframe by using '~:mtcars'.

```{ruby nice_table}
R.library('kableExtra')
outputs (~:mtcars).kable.kable_styling
```

### Including Ruby files

R is a language that was created to be easy and fast for statisticians to use.  As far
as I know (and please correct me if you think otherwise), tt was not a
language to be used for developing large systems.  Of course, there are large systems and
libraries in R, but the focus of the language is for developing statistical models and
distribute that to peers.

Ruby on the other hand, is a language for large software development.  Systems written in
Ruby will have dozens, hundreds or even thousands of files.  In order to document a 
large system with
literate programming we cannot expect the developer to add all the files in a single '.Rmd'
file.  gKnit provides the 'include' chunk engine to include a Ruby file as if it had being
typed in the '.Rmd' file.

To include a file, the following chunk should be created, where <filename> is the name of
the file to be include and where the extension, if it is '.rb', does not need to be added.
If the 'relative' option is not included, then it is treated as TRUE.  When 'relative' is
true, 'require_relative' semantics is used to load the file, when false, Ruby's \$LOAD_PATH
is searched to find the file and it is 'require'd.

````
```{include <filename>, relative = <TRUE/FALSE>}`r ''`
```
````

Here we include file 'model.rb' which is in the same directory of this blog.  
This code uses R 'caret' package to split a dataset in a train and test sets.
The 'caret' package is a very important a useful package for doing Data Analysis,
it has hundreds of functions for all steps of the Data Analysis workflow.  To
just split a dataset it is using the proverbial cannon to kill the fly.  We use
it here only to show that integrating Ruby and R and using even a very comples
package as 'caret' is trivial with Galaaz.

A word of advice: the 'caret' package has lots of dependencies and installing
it in a Linux system is a time consuming operation.  Method 'R.install_and_loads'
will install the package if it is not already installed and can take a while.

````
```{include model}`r ''`
```
````

```{include model}
```

```{ruby model_partition}
mtcars = ~:mtcars
model = Model.new(mtcars, percent_train: 0.8)
model.partition(:mpg)
puts model.train.head
puts model.test.head
```

### Documenting Gems

gKnit also allows developers to document and load files that are not in the same directory
of the '.Rmd' file.  When using 'relative = FALSE' in a chunk header, gKnit will look for the
file in Ruby's \$LOAD_PATH and load it if found.

Here is an example of loading the 'find.rb' file from TruffleRuby.

````
```{include find, relative = FALSE}`r ''`
```
````

```{include find, relative = FALSE}
```

## Converting to PDF

One of the beauties of knitr is that the same input can be converted to many different outputs.
One very useful format, is, of course, PDF.  In order to converted an R markdown file to PDF
it is necessary to have LaTeX installed on the system.  We will not explain here how to
install LaTeX as there are plenty of documents on the web showing how to proceed.

gKnit comes with a simple LaTeX style file for gknitting this blog as a PDF document.  Here is
the Yaml header to generate this blog in PDF format instead of HTML:

```
---
title: "gKnit - Ruby and R Knitting with Galaaz in GraalVM"
author: "Rodrigo Botafogo"
tags: [Galaaz, Ruby, R, TruffleRuby, FastR, GraalVM, knitr, gknit]
date: "29 October 2018"
output:
  pdf_document:
    includes:
      in_header: ["../../sty/galaaz.sty"]
    number_sections: yes
---
```

# Conclusion

One of the promises of GraalVM is that users/developers will be able to use the best tool
for their task at hand, independently of the programming language the tool was written.  Galaaz
and gKnit are not trivial implementations atop the GraalVM and Truffle interop messages;
however, the time and effort it took to wrap Ruby over R - Galaaz - (not finished yet) or to
wrap Knitr with gKnit is a fraction of a fraction of a fraction of the time require to
implement the original tools.  Trying to reimplement all R packages in Ruby would require the
same effort it is taking Python to implement NumPy, Panda and all supporting libraries and it
is unlikely that this effort would ever be done.  GraalVM has allowed Ruby to profit "almost
for free" from this huge set of libraries and tools that make R one of the most used
languages for data analysis and machine learning.

More interesting though than being able to wrap the R libraries with Ruby, is that Ruby adds
value to R, by allowing developers to use powerful and modern constructs for code reuse that
are not the strong points of R.  As shown in this blog, R and Ruby can easily communicate
and R can be structured in classes and modules in a way that greatly expands its power and
readability.

# Installing gKnit

## Prerequisites

* GraalVM (>= rc8)
* TruffleRuby
* FastR

The following R packages will be automatically installed when necessary, but could be installed prior
to using gKnit if desired:

* ggplot2
* gridExtra
* knitr

Installation of R packages requires a development environment and can be time consuming.  In Linux,
the gnu compiler and tools should be enough.  I am not sure what is needed on the Mac.

## Preparation

* gem install galaaz

## Usage

* gknit \<filename\>
