---
title: "gKnit - Ruby and R Knitting with Galaaz"
author: "Rodrigo Botafogo"
tags: [Galaaz, Ruby, R, TruffleRuby, FastR, GraalVM, knitr]
date: "19 October 2018"
output:
  html_document:
    self_contained: true
    keep_md: true
---

```{r setup, echo=FALSE}

```

# Introduction

The idea of "literate programming" was first introduced by Donald Knuth in the 1980's.
The main intention of this approach was to develop code and at the same time document it
is as understandably as possible to humans.  According to Knuth "The practitioner of 
literate programming can be regarded as an essayist, whose main concern is with exposition 
and excellence of style."

The idea of literate programming envolved into the idea of reproducible research, in which
all the data, information and documentation needed to reproduce a research is included in a
single document that can be distributed to peers that can rerun the research obtaining
the same results.

The R community has put a great deal of effort in reproducible research.  In 2002, Sweave was
introduced and it allowed mixing R code with Latex generating hight quality PDF documents.  Those
documents could include the code, the result of executing the code, graphics and text.  This
contained the whole narrative to reproduce the research.  But Sweave had many problems and in
2012, Knitr, developed by Yihui Xie from RStudio was released, solving many of the long lasting
problems from Sweave and including in one single package many extensions and add-on packages that
were necessary for Sweave.  With Knitr, R markdown was also developed, an extension the the
Markdown format.  With R markdown and Knitr it is possible to generate reports in a multitude
of formats such as HTML, markdown, Latex, PDF, dvi, etc.  R markdown also allows the use of
multiple programming languages in the same document.  However, other than R and Python (with
the reticulate package), each chunk of code in any other language will execute in a new
process, so memory and variables cannot be shared between chunks.

In the Python community, the same effort to have code and text in an integrated environment
started also on the first decade of 2000 and in 2006 iPython 0.7.2 was released.  In 2014,
Fernando Pérez, spun off project Jupyter from iPython creating a web-based interactive
computation environment.  Jupyter can now be used with many languages, including Ruby with the
iruby gem (https://github.com/SciRuby/iruby).  I am not sure if multiple languages can be used
in a Jupyter notebook.

# gKnitting a Document

This document describes gKnit.  gKnit used Knitr and R markdown to knit a document and output
it in any of the
available formats for R markdown.  The only difference between gKnit and normal Knitr documents
is that gKnit runs atop of GraalVM, and Galaaz (an integration library between Ruby and R).
Another blog post on Galaaz and its integration with ggplot2 can be found at:
https://towardsdatascience.com/ruby-plotting-with-galaaz-an-example-of-tightly-coupling-ruby-and-r-in-graalvm-520b69e21021.  With Galaaz gKnit can knit documents in Ruby and R and both
Ruby and R execute on the same process
sharing variables and memory and variables, classes, etc. will be preserved between chunks of
code.

This is not a blog post on rmarkdown, and the intereste user is directed to
https://rmarkdown.rstudio.com/ or
https://bookdown.org/yihui/rmarkdown/ for detailed information on its capabilities and use.
Some of the features available for R code
chunks are not yet available for Ruby code.  But we are working for removing any limitations.
Here, we will describe quickly the main aspects of R markdown, so the user can start gKnit simple
documents quickly.

## The Yaml header

An R markdown document should start with a Yaml header.  This document has the following header that
when gKitted will generate and HTML document.

```
---
title: "gKnit - Ruby and R Knitting with Galaaz"
author: "Rodrigo Botafogo"
tags: [Galaaz, Ruby, R, TruffleRuby, FastR, GraalVM, knitr]
date: "19 October 2018"
output:
  html_document:
    keep_md: true
includes:
      in_header: ../../bin/header.html
      after_body: ../../bin/footer.html    
---
```

For more information on the options in the Yaml header, check https://bookdown.org/yihui/rmarkdown/html-document.html.

## R Markdown formatting

Document formating can be done with simple markups such as:

### Headers

```
# Header 1

## Header 2

### Header 3

```

### Lists

```
Unordered lists:

* Item 1
* Item 2
    + Item 2a
    + Item 2b
```

```
Ordered Lists

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```

Please, go to https://rmarkdown.rstudio.com/authoring_basics.html, for more R markdown formating.

## Code Chunks

### R chunks

Running and executing Ruby and R code is actually what really interests us is this blog.  Doing so
is very simple with code chunks.

In order to add R code to a text, a code chunk is added with the following markup:

So, for instance, here we create an R code chunk setting a variable to a vector:

````
```{r, eval=TRUE}`r ''`
r_vec <- c(1, 2, 3)
```
````

```{r var_setting}
r_vec <- c(1, 2, 3)
# Using an example vector "arg"
arg = c(1, 2, 3, 4, 5)
redundent_sum <- function(...) {
  Reduce(sum, as.list(...))
}
redundent_sum(arg)
```


### Ruby chunks

```{ruby split_data}
$a = [1, 2, 3]
$vec = R.c(1, 2, 3)
$vec2 = R.c(10, 20, 30)
$b = "R$ 250.000"
$c = "class Array"
```

I'm now testing the ruby engine with more comples code in it.
```{ruby split2}
puts $a
puts $vec * $vec2
```
we need to check what the result is gona be

```{ruby ruby_high, eval=FALSE}
a = String.new("hello there")
b = [1, 2, 3]
puts a
puts b
```

### Accessing R from Ruby


### Inline Ruby code

This is some text with inline Ruby
```{rb puts $b}
```
is this lost will this continue?

This is also possible
```{rb puts $a; puts $b}
```
what comes out??

And what about this?

```{rb puts "# #{$c}"}
```
Any problems surround it with text?

### Including Ruby files

```{include model}
```


# Installing gKnit

## Prerequisites

* GraalVM (>= rc7)
* TruffleRuby
* FastR

The following R packages will be automatically installed when necessary, but could be installed prior
to using gKnit if desired:

* ggplot2
* gridExtra
* knitr

Installation of R packages requires a development environment and can be time consuming.  In Linux,
the gnu compiler and tools should be enough.  I am not sure what is needed on the Mac.

## Preparation

* gem install galaaz

## Usage

* gknit [filename]